---
title: "Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies"
author: "AJ Perez-Luque; G. Gea-Izquierdo; Zamora, R. ..."
output:
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    highlight: null
    reference_docx: ../templates/template_ms.docx
bibliography: ../refs/references.bib
csl: ../refs/ecology.csl
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(here)
library(pander)
library(kableExtra)
library(knitr)
```

```{r setup, echo=FALSE}
show_text <- FALSE
```

# Abstract 
* Global change challenges forest ecosystems particularly in species-specific dry locations where land-use legacies have modified species composition and the forest structure as a result of intensive human pressure in the past. We analyzed forest resilience of adult trees to drought and climate change of relict Mediterranean *Quercus pyrenaica* Willd. populations in the Southern Iberian Peninsula. Forest resilience was analyzed using both dendroecological estimates of growth and MODIS satellite images as proxies of secondary and primary growth respectively.

* The studied relict populations showed expressions of management, likely reflecting the origin of the current forest structure. Both resilience indices of primary and secondary growth expressed that trees showed high resilience both to recent drought events and also to long-term climatic changes (i.e. warming), despite the ‘a priori’ vulnerability that we were expecting them to express in respone climate change on relict stands resulting from intensive land-use in history at their current rear-edge. Furthermore, the coldest stands show a positive growth trend since the end of the 1970s matching the simultaneous increase in temperature observed.

* Resilience indices showed ecological coherence along a gradient of drought severity in time. In addition, oak resilience to drought followed a dryness gradient among the three populations studied: Northern populations thriving under a drier climate were less resilient to drought both in primary and secondary growth than Southern populations where lower altitudes were also less resilient than those at high-altitudes.

* Despite being considered relict stands at the dry rear-edge for the species, they showed high resilience in response to drought. Yet, the high sensitivity to water availability expressed and positive growth trends only in the coldest and more humid stands suggest divergent response to a changing climate along very short climatic gradients and suggest that dynamics (i.e. not only adult performance but also regeneration) of those stands located at the lowest elevations, hence the altitudinal rear-edge, need to be monitored under future expected drier and warmer climatic scenarios and, particularly, if high management pressure remains. 

# Introduction

Drought is a natural phenomenon occurring when water availability is significantly below-normal levels [@Dai2011]. It is one of the most frequent and widespread climatic extremes affecting ecological systems across every climatic zone worldwide [@Dai2011; @Ummenhofer2017]. Despite drought is considered a feature of Mediterranean climate [@Lionello2012], an increase in the frequency and severity of drought events has been recorded in the last decades [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported [@Spinoni2017b]. This is especially important, since the climate change projections for the Mediterranean area forecast that climate extreme events, like droughts, will become more frequent and severe [@IPCC2013; @Trenberth2014; @Spinoni2017a].

Drought causes significant changes in ecosystem productivity and water dynamics that can severely affect forests through changes in plant physiology, phenology and carbon allocation [@Allen2010; @Ummenhofer2017]. The assessment of the drought impacts on ecosystems has gained attention in the last decades [@Allen2010; @Clark2016], and the functional response of vegetation to drought has been analyzed at global [*e.g.*, @VicenteSerrano2013; @Norman2016] and at local scales [see @MartinezVilalta2016 for a revision]. 

Ecosystem responses to drought are influenced by other drivers of environmental changes [@Fischer2006; @Oliver2014; @Penuelas2017]. In a global-change context, the impacts of drought on vegetation need to be evaluate considering the interactions with other drivers of change [@Clavero2011; @DoblasMiranda2017]. This is especially relevant for areas with a long history of landscape modification, such us Mediterranean region, where land-use change plays a significant role [@Navarro2013; @Ameztegui2016], considering also that the interactions of these two factors, namely drought and land-use change, are crucial for Mediterranean forests [@DoblasMiranda2017].

The impacts of drought are also especially relevant for populations located in the rear edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions often corresponds to the species-limits and the populations are likely to be more sensitive to small variations on climate [@Hampe2005]. The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010]. It becomes even more important for relict populations driven by historical land-uses, which can harbour high levels of intraspecific genetic diversity [@Valbuena2010].

Several studies analyzed the drought effects on Mediterranean tree species representing the southernmost limit of their distribution
[*e.g.*, @SanchezSalguero2012; @Linares2014; @Dorado2017b; @Caminero2018]. Some of these studies focused on the resilience to drought of rear-edge populations [*e.g.*, @Herrero2014]. Resilience is the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance [@Holling1973; @Hodgson2015]. The assessment of resilience to climate disturbances provides critical information about the capacity of the forests to maintain their structure and to continue providing valuable ecosystem services. This is particularly relevant for populations located in the rear-edge of their current distribution, especially when they are located in Mediterranean mountains.

Here we assessed the resilience to drought of rear-edge populations of a tree Mediterranean species combining several approaches: remote sensing and dendroecologial estimates of growth. Remote sensing can be used for studying droughts impacts on ecosystems [*e.g.*, @Deshayes2006; @Zhang2013; @AghaKouchak2015; @McDowell2015; @Norman2016]. Vegetation indices obtained from satellite, such us EVI (Enhanced Vegetation Index) are used as a proxy for assessing vegetation functioning and aboveground net primary production [@Huete2002], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013; @Vicca2016]. Drought impacts on vegetation using remote sensing have been analyzed at different scales. For instance, a global-scale study showed a stronger influence of the drought for drier biomes [@VicenteSerrano2013]. For Mediterranean area, studies at regional and local scales found a large spatial and seasonal variability of the effects of drought on vegetation [@VicenteSerrano2007; @Gouveia2017]. 

Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality [@Fritts1976; @Dobbertin2005; @Bhuyan2017] and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976]. The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological or elevational gradients [*e.g.*, @Dorman2013; @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018]

In this work we used remote sensing information and dendroecological methods to evaluate the drought impacts in both canopy greenness (as a proxy to primary growth) and tree-radial growth of a Mediterranean tree species (*Quercus pyrenaica* Willd.) located in Sierra Nevada, which is considered a rear-edge of species distribution and species performance is considered to be threatened. We also assess the resilience of these forests both to several extreme drought episodes and to climate change in the long-term in the last decades.   
Our hypothesis is that these relict populations driven by historical land-use at their climatic (either altitudinal or latitudinal) rear-edge are particularly vulnerable to climate change, and hence they will show low-values of resilience after several extreme droughts. Our specifics aims were: (*i*) To quantify how two recent extreme drought events influenced primary and secondary growth of *Q. pyrenaica* forests in their rear edge; (*ii*) to analyze the long-term resilience of these forests to extreme drought events using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of the distribution of this species to detect vulnerability to climate change. In addition, within the region of study we were interested to assess whether the effect of aspect and microclimatic conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.


```{r echo=FALSE, results='asis', eval=show_text}
cat("

In the Iberian Peninsula, major drought episodies were recorded in 1943, 1981, 1995, 2000 and 2005 [@VicenteSerrano2014; @Guerreiro2017]. The 2004/2005 and 2011/2012 hydrological years are considered two of the worst drought periods recorded in the Iberian Peninsula, particularly in the southern sector [@GarciaHerrera2007; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017]. These events were extreme in both its magnitude and spatial extent [@Gouveia2014].


These climatic alterations are likely to have important consequences for tree species dynamics at local and regional scales ( Peñuelas and Boada 2003 , Van Mantgem et al. 2009 , Matías and Jump 2015 ).









### Sequías 
> notas sequia
> 
> * the drought events have been longest and most severe in the period 1991-2010 for mediterranean are of Southern Europe  
> * Seasonally, drought frequency is projected to increase everywhere in Europe for both scenarios in spring and summer, especially over southern Europe, and less intensely in autumn [@Spinoni2017a]
> Aumento de la frecuencia de sequías en spring and summer desde 1950 hasta 2014 [@Spinoni2017b]
> Both for frequency and severity, the evolution towards drier conditions is more relevant in the last three decades over Mediterranean area in summer, 
> an increase in the drought severity in the Iberian Peninsula has been observed in the last decades [@VicenteSerrano2014]. 

Althought several works have reported these two years as some of the worst drought events, we characterised the drought at several spatio-temporal scales in the study area. From a long-term perspective, we compare the accumulated monthly precipitation at a meteorological station (Granada, Base Aérea) during the hydrological years 2004-2005 and 2011-2012 with the average of accumulated monthly precipitation for the period 1950-2015.

### Extreme sequias y ecologia 


No solo efectos aislados, sino también en conjunción con otros factores, sobre todo en el mediterráneo donde convergen muchos de los factores que pueden interaccionar: 

* La sequía es un factor crucial a tener en cuenta, ya que es además de los efectos que puede tener de forma aislada, se ha visto que además presenta muchas interacciones con otros factores, siendo por tanto un factor crucial (`$IMPROVE$` ver [@DoblasMiranda2017]):
* Some interactions alter the effects of a single factor, as drought enhances or decreases the effects of atmospheric components on plant ecophysiology
* Drought and land use changes, among others, alter water resources and lead to land degradation, vegetation regeneration decline, and expansion of forest diseases.
* Climate change, and especially drought, emerges as a crucial factor in most of the reviewed interactions and therefore it should be considered when it comes to designing and applying international management policies
* Drought should be considered when designing and applying management policies.


Además la sequía se espera que tengan ...Droughts are most likely to have the largest and most long-lasting impacts globally due to large indirect and lagged impacts and long recovery especially for forest ecosystems (ver  18 en @Ummenhofer2017). Así por ejemplo se ha visto que todos los biomas presentan una vulnerabilidad similar ...
En una revision sobre la vulnerabilidad del sistema de trasnporte en plantas al embolismo inducido por sequía, ha mostrado que una convergencia de la vulnerabilidad de los bosques a la sequía, mostrando que todos los biomas son igualmente vulnerables a los fallos hidráulicos independientemente del régimen de precipitación (Choat et al. 2012, doi:10.1038/nature11688)

")
```

# Materials and methods 

## Tree species and study site 

Pyrenean oak (*Quercus pyrenaica* Willd.) forests extend through south-western France and the Iberian Peninsula reaching its southern limit in mountains areas of northern Morocco [@Franco1990]. In the Iberian Peninsula these forests live on siliceous soils, or soils poor in basic ions [@Vilches2014] under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate [@RivasMartinez2002]. *Q. pyrenaica* requires over 650 mm of annual precipitation and a summer minimal precipitation between 100 and 200 mm [@MartinezParras1982; @GarciaJimenez2009], with summer rainfall being a key factor in the distribution of the species, which can hence be considered to have low drought tolerance compared to genuine evergreen Mediterranean taxa [@Gavilan2007; @delRio2007].

The forests of this species reaches their southernmost European limit in Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations between 860 m and 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August). There are eight Pyrenean oak patches (2400 Has) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys. Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010] and these populations are considered as a rear edge of the habitat distribution, which is important in determining habitat responses to expected climate change [@Hampe2005]. For this reason, the populations of Pyrenean oak forests at Sierra Nevada are considered relict forests [@Melendo2000; @Vivero2000], having these forests undergone intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2010; @Valbuena2017]. Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" [@Vivero2000] and they are expected to suffer the impact of climate change, reducting its suitable habitats in the rear-edge due to their greater hydric requirements compared to other more drought-tolerant Mediterranean evergreen oak and pine species [@GeaIzquierdo2013].

## Drought episodes

During the second half of the twentieth century the Iberian Peninsula suffered several extreme drought episodes (*e.g*. 1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014; @Guerreiro2017].The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2014; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017] and they were characterized as extreme drought in our climatic data (Figures S1-S2; Table S3). We focused on these two drought events because they were those included in the period where there is availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events and analyze forest resilience to drought in a longer term (see Table S3). The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Spinoni2015. We used SPEI-12 months scale for all 0.5º grid cells covering Sierra Nevada obtained from Global SPEI Database (http://spei.csic.es/database.html). A severe drought event starts in the month when SPEI falls below the threshold of -1.28 [@Spinoni2017a; @Pascoa2017]. A drought event is considered only when SPEI values are below the threshold for at least two consecutive months [*e.g.* @Spinoni2015; @Spinoni2017a]. Several indicators were computed for each drought event: ***duration*** as the number of consecutive months with the SPEI lower than a certain threshold; ***severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event; the ***intensity*** and the ***Lowest SPEI*** refer to the mean and lowest value of SPEI respectively during the drought event duration.  


```{r echo=FALSE, results='asis', eval=show_text}
cat("
* Ver ejemplo Gazol et al 2017 Plan Ecology
* un apartado llamado Drought episodes (similar a esto https://www.nature.com/articles/srep28269)
")
```

## Greenness data 

Vegetation indices derived from remote sensing information have been widely used to derive descriptors of ecosystem functioning [*e.g* @AlcarazSegura2009] because they are linked with primary productivity [@Ruimy1994]. Here, vegetation greenness of *Q. pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the *Moderate Resolution Imaging Spectroradiometer* (MODIS) sensor [@Didan2015]. MODIS EVI Data (Collection 6) were obtanied using Google Earth Engine platform for the period 2000 - 2016. EVI data consits of 16-day maximun value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m. We selected the pixels covering the distribution of *Q. pyrenaica* forests in Sierra Nevada (*n* = 928 pixels). Then, a data filtering was applied to select EVI valid values. For this purpose, complementary information accompanying the EVI data was used (VI Usefulness Indices and quality flags). Values affected by high content of aerosols, clouds, snow and shadows, were filtered out following recommendations for mountain regions [@Reyes2015]. 

```{r echo=FALSE, results='asis', eval=show_text}
cat("
We used EVI instead of NDVI (*Normalized Difference Vegetation Index*; another of the most common greenness vegetation indices) because EVI is more sensitive to changes in high-biomass areas (a serious shortcoming of NDVI); EVI reduces the influence of atmospheric conditions on vegetation index values, and EVI corrects for canopy background signals [@Huete2002; @Krapivin2015; @Cabello2012]. 

")
```

```{r echo=FALSE, results='asis', eval=show_text}
cat("
Each 1 × 1 km2 16-day composite EVI value is considered valid when (a) EVI data is produced—'MODLAND_QA' equals 0 (good quality) or 1 (check other QA), (b) VI usefulness is between 0 and 11, (c) clouds are absent—'adjacent cloud detected' (0), 'mixed clouds' (0) and 'possible shadow' (0), and (d) aerosol content is low or average—'aerosol quantity' (1 or 2). Note that 'MODLAND_QA' checks whether EVI is produced or not, and if produced, its quality is good or whether other quality flags should also be checked. Besides, VI usefulness indices between 0 to 11 essentially include all EVI data. Thus, these two conditions serve as additional checks.



and then a quality assessment was carried out to filter the ... [@Reyes2015] 

    * According to @Reyes2015 we must consider the shadow in the mountain, but we can discard the filter of adjacent clouds. On the other hand, the use of EVI mean is highly stable under the use of any filter [@Reyes2015]


The presence of clouds (adjacent clouds, mixed clouds and shadows) 'obscures' the surface in a radiometric sense, thus corrupting inferred EVI values. In addition, two types of aerosol loadings typically corrupt EVI—climatology and high aerosols. Use of aerosol climatology 


The EVI data are geometrically and atmospherically corrected and include information about the quality ass.... 

These data are geometrically and atmospherically corrected, and include an index of data quality (reliability, which range from 0 – good quality data – to 4 – raw data or absent for different reasons) based on the environmental conditions in which the data was recorded

We first used the Quality Assesment (QA band) information of this product to filter out those values affected by high content of aerosols, clouds, shadows, snow or water; and then a quality assessment was carried out to filter the ... [@Reyes2015] 

:red_circle: reescribir esto de la calidad. Ver Samanta et al 2012 y como describe el proceso de calidad

")
```

As a surrogate of mean annual primary production, annual mean EVI ($EVI_{mean}$) value was computed for each pixel for the period 2000 - 2016. To explore the effect of drought on greenness, the EVI standardized anomaly ($EVI~sa$) was computed pixel-by-pixel, since it minimizes biases in the evaluation of anomalies, providing more information about the magnitude of the anomalies [@Samanta2012; @Gao2016]. For each pixel, an annual EVI value was obtained by averaging EVI valid values. Then, the standardized anomaly was computed as: 

$$\mathrm{EVI_{sa,\mathit{i}}}= \frac{\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}}}{\sigma_{\mathrm{ref}}}$$
where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except $i$ year); and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period. Each pixel was categorized according the EVI standardized anomalies as "greening" (EVI standardized anomalies greater than + 1), "browning" (EVI standardized anomalies less than - 1) or "no changes" (EVI standardized anomalies between − 1 and  + 1) [@Samanta2010; @Samanta2012]. 

Rather than other vegetation indices like the NDVI, $EVI_{mean}$ was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual ($r$ = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88). For exploratory purposes, temporal trends of $EVI_{mean}$ were explored at a pixel scale. The nonparametric Mann–Kendall test (Kendall, 1975; Mann, 1945) was applied using the `trend` package in R [@Thorsten2017]. The alpha level for the test was set to 0.05. 

```{r echo=FALSE, results='asis', eval=show_text}
cat("
(:red_circle: Hemos seleccionado EVI medio, además de por los consejos que me ha dicho Domingo, porque he comprobado que existe una correlación entre el evi medio y el evi estacional, sobre todo el de verano. Ver esto: [https://github.com/ajpelu/qpyr_modis_resilience/blob/master/analysis/prepare_modis_qa.md](https://github.com/ajpelu/qpyr_modis_resilience/blob/master/analysis/prepare_modis_qa.md). Además presenta alta correlaciones significativas con el EVI de verano: 0.88; de primavera: 0.76 y anual: 0.81)


> Procedimiento de Filtrado de datos (ver [https://github.com/ajpelu/qpyr_modis_resilience/blob/master/analysis/prepare_modis_qa.md](https://github.com/ajpelu/qpyr_modis_resilience/blob/master/analysis/prepare_modis_qa.md))

* Información contenida en banda QA. 

    * Nos quedamos con pixeles marcados como Good Data (57.89 %)
    * Filter out los marcados como Snow/Ice y/o Cloudy (2.57 + 7.08 = 9.65 %)
    * Pixeles marcados como Marginal Data (32.33 %) (ver siguiente paso)

* Explorar distribución temporal y analizar banda QA Detailed y llevar a cabo un filtrado siguiendo las especificaciones de @Reyes2015.  

    * Vemos los composites marcados con Aerosoles, Adjacent cluods, y Shadow.  
    * According to @Reyes2015 we must consider the shadow in the mountain, but we can discard the filter of adjacent clouds. On the other hand, the use of EVI mean is highly stable under the use of any filter [@Reyes2015]
    
* Finalmente nos hemos quedado con las siguientes cifras. De un total de 360064 images composites for the study zone were downloaded (928 x 20 x 1 + 928 x 23 x 16 = 360064), tras el filtrado, nos quedamos con 286825 (79.65 %)

")
```

## Field sampling and dendrochronological methods 

### Tree sampling 
Tree sampling was carried out during autumn 2016. Trees were sampled at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at northern aspect; and Cáñar (CA), a humid but warm site located at southern aspect (Figure 1; Table 1). For the southern site two elevations were sampled: CA-Low and CA-High, which constitutes the current altitudinal tree-line in the sampled slope. All the sites were oak monospecific and representatives of the population clusters identified for the species in this mountain range [@PerezLuque2015]. In each site between 15 and 20 dominant trees were randomly selected. Two cores of 5 mm of diameter were taken from each tree at breast heigth (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured using a girth tape and a Vertex IV ultrasonic hypsometer (Haglöf, Sweden) respectively. Stand competition of target trees was assessed by recording distance, azimuth, DBH, species and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot of 10 m radius. Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$. Then, non-parametric Kruskal-Wallis rank sum tests [@Sokal1995] were used to analyse differences between sites for height, DBH and competition indices. When significant differences were observed, we run multiple comparisons using the Dunn's-test [@Dunn1964] with Bonferroni adjustment to correct significance values. These analysis were performed uisng the `PMCMR` package in R [@Pohlert2014]. 

### Dendroecological analyses 
Tree cores were air dried, glued onto wooden mounts and sanded. Annual radial growth (ring width, RW) was measured with a LINTAB measuring device (Rinntech, Heidelberg, Germany) coupled to a stereomicroscope, with an accuracy of 0.001 mm. Individual ring series were first visually and statistically cross-dated with TSAP software (Rinntech, Heidelberg, Germany), using the statistics Gleichläufigkeit (GLK), t-value and the crossdating index (CDI). Cross-dating validation was finally done using COFECHA software [@Holmes1983].

The growth trends were analyzed at different time scales. To study the response of growth to the inter-annual variability of climate (short-term response), pre-whitened residual chronologies (RWI) were used. They were calculated from ratios between raw growth measurements and individual cubic splines with a 50 % frequency cutoff at 30 years [@Fritts1976]. Tree-ring width series were standardized and detrended using the `dplR` [@Bunn2008; @Bunn2010] package. Residual site chronologies were obtained computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Cook1990]. The statistical quality of each chronology was checked via the expressed population signal. A threshold value of EPS > 0.85 was used to determine the cutoff year of the time span that could be considered reliable [@Wigley1984].

The long-term growth response was analyzed using basal area increment (hereafter BAI, $\mathrm{cm^2 \cdot year^{-1}}$). Theoretically, BAI represents a more accurate indicator of growth than ring-width, since it removes variation in growth attributable to increasing stem circumference after 30-40 years of juvenile increasing growth [@Biondi2008]. Raw ring-widths and measured DBH were used to compute BAI by subtracting twice the annual ring width from the annual diameter, starting from the measured diameter outside the bark [@Piovesan2008]. We used the following equation:

$$\mathrm{BAI} = \pi(r^2_{t}-r^2_{t-1})$$
where $r$ is the radius of the tree and $t$ is the year of tree-ring formation. For each individual tree, a mean BAI serie was calculated. Then, mean site BAI chronologies were obtained by averaging individual tree BAI time series. 

Both RWI and BAI chronologies were compared to different climatic indices to analyse the low and the high-frequency response to climate of trees. Statistical descriptive parameters, including the mean, standard deviation, first-order autocorrelation of raw series, the mean sensitivity (a measure of the year-to-year variability) and the mean correlation between individual series of residual ring-width indices, were also calculated for each site chronology [@Fritts1976]. 

### Climate and growth 

Climate data were obtained from the European Daily High-Resolution Observational Gridded Dataset (E-OBS v16) [@Haylock2008]. Monthly precipitation and minimum and maximum temperatures were obtained at a 0.25 x 0.25 º resolution for the 1950-2016 period. We selected grid cells covering each sampled sites. Data were downloaded and preprocessed using the climate4R bundle (http://www.meteo.unican.es/en/climate4R) [@Cofino2018; @Frias2018]. The Standardized Precipitation-Evapotranspiration Index (SPEI), a multiscalar drought index that incorporates both precipitation and temperature [@VicenteSerrano2010; @Begueria2014], was also used to characterize the drought conditions for the period 1961-2014. SPEI values with a temporal scale of 6 months and a spatial resolution of 1.1 km, were obtanied from the Drought indices dataset (http://monitordesequia.csic.es/), a high resolution database of drought indices for Spain [@VicenteSerrano2017]. 

Relationships between residual site chronologies and climatic variables were assessed ussing bootstrapped Pearson's correlations estimated using the `treeclim` package [@Zang2015]. The bootstrapped confidence intervals were used to estimate significance (p < 0.05) of the correlation coefficients.

```{r echo=FALSE, results='asis', eval=show_text}
cat("
SPEI values between 1 and −1 are considered normal, whereas values <−1 indicate drought and values <−1.5 indicate severe drought47. We defined a drought as a period where SPEI12 decreased to ≤−1.5.
")
```


### Disturbance analyses 
Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (supressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (e.g. logging, neighbor mortality) in the past. By using a temporal window of reasonable length this method is intended to filter out the response to short-term changes in temperature and precipitation and minimizes the long-term growth trends [@Nowacki1997; @Fraver2005]. 

```{r echo=FALSE, results='asis', eval=show_text}
cat("
This approach minimizes the long-term growth trends and interannual growth variations usually driven by climate, while enhances decadal abrupt and sustained radial-growth changes characteristic of forest disturbances.
")
```

Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes:

$$\mathrm{\% GC} = \left [ \frac{\left ( M1 - M2 \right )}{M2}\right] \times 100 $$
where $M1$ is preceding 10-year median and $M2$ is subsequent 10-year median. Medians were used since they are more robust estimator of central tendency than means [@Rubino2004; @Camarero2011]. Site disturbance chronologies were constructed by averaging the individual disturbances series annually. To separate growth peaks produced by disturbance events and expressing stand-wise disturbances from those by climate, we considered a threshold of 50 % of GC and more than 50% of the individual trees displaying the same growth changes was considered a stand-wise disturbance. 

```{r echo=FALSE, results='asis', eval=show_text}
cat("
Disturbance chronologies were built using tree-ring width to identify historical abrupt positive (releases) or negative (suppressions) changes in growth in each species (Nowacki and Abrams, 1997). This approach minimizes the long-term growth trends and interannual growth variations usually driven by climate, while enhances decadal abrupt and sustained radial-growth changes characteristic of forest disturbances.
")
```

```{r echo=FALSE, results='asis', eval=show_text}
cat("
* Para obtener los datos de SPEI, hemos ido bajado los datos desde http://sac.csic.es/spei/map/maps.html 
* sanjuan: 37.133, -3.382 = spei_-3.25_37.25.csv 
* canar: 36.957, -3.4284 = spei_-3.25_36.75.csv

OJO ver http://monitordesequia.csic.es/
")
```

```{r echo=FALSE, results='asis', eval=show_text}
cat("
To obtain ring-width series of each individual and species (species chronology) the cross-dated ring-width series were standardized and detrended using the dplR library (Bunn 2010) in the R statistical language (R Development Core Team 2015). First, we fitted a spline through the ring-width series of each measured radius. We used a spline where the frequency response is 0.5 at a wavelength of 0.67 multiplied by the length of the series in years (Bunn 2010). Then, the ring-width data were divided by the fitted lines to obtain residual ring-width indices (RWI) that were subjected to autoregressive modelling (pre-whitening) and then averaged for each year using a biweight robust estimation of the mean (Fritts 1976). Calculating RWIs using subtraction instead of division rendered similar results as those here presented. Thus, we obtained mean residual chronologies for each species of prewhitened growth indices.

 * BAI represents a more accurate indicator of growth than ring-width, since it reduces the variation caused by adding volume to a circular stem 
 * BAI is a more meaningful indicator of growth than ring width because it removes variation in growth attributale to increasing circumference. 
 * BAI is more closely related to biomass increment 
 * BAI removes variation in growth attributatle to increasing stem circumference and captures changes in growth better than linear measures as tree-ring width [@Biondi2008] 
")
```

```{r echo=FALSE, results='asis', eval=show_text}
cat("
 To explore similarity within locality, each site chronology was smoothed using centred moving averages with different window sizes, and then Pearson's correlation coefficient between the two chronologies of the same locality (higher and lower elevation) were calculated. Significance was tested using 1000 boostrap replicates and with 95 % confidence intervals built using the R packgae `boot` [@Canty2016]
")
```

In addition, the forest history of San Juan and Cáñar sites was inferred from a detailed analysis of historical land-use changes. For this purpose, an exhaustive review of historical documents was done to compile information about socio-economical activities affecting forest: land-uses, forest managment actions, mining, wildfires, etc. Several documentary sources were reviewed: historical documents and maps [*e.g.* @Catastro1752; @Ministerio1943; @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information about recent wildfires events (Regional Ministry for the Environment of the Government of Andalusia) and forest management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2014]; traditional irrigation ditches [@MartinMontanes2015; @RuizRuiz2017] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@Bonet2014_conama; @JimenezOlivencia2015; @MorenoLlorca2016].


## Resilience 

To evaluate the effects of drougth events on greeennes and tree growth (BAI) we used resilience indices proposed by @Lloret2011: resilience (*Rs*), resistance (*Rt*), and recovery (*Rc*). The resistance index (*Rt*) quantifies the severity of the impact of the disturbance in the year it occurred. It is estimated as the ratio between the performance during and before the disturbance (drought) event:

Resistance (*Rt*) = Drought / Predrought

The Recovery index (*Rc*) is the ability to recover from disturbance relative to its severity, and it is estimated as the ratio between performance after and during disturbance: 

Recovery (*Rc*) = Postdrought / Drought

The Resilience index (*Rs*) is the capacity to reach pre-disturbance performance levels, and it is estimated as the ratio between the performance after and before disturbance: 

Resilience (*Rs*) = Postdrought / Predrought

```{r echo=FALSE, results='asis', eval=show_text}
cat("
The Relative Resilience (*RRs*) is the resilience weighted by the severity of the disturbance, and it is estimated as: 

Relative Resilience (*RRs*) = (Postdrought - Drought) / Predrought
")
```

The values of these indices were computed for tree growth (BAI) and greenness (EVI mean) during each drought event. 2005 and 2012 were considered as single drought events. The predrought and postdrought values of each target variable (*i.e.*: BAI or EVI) were computed as the mean value during a period of three years before and after the drought event respectively. A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S6) and this length was used in other studies [*e.g.* @Gazol2017]. To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1900 and compared them with drought severity (Figure 6, Table S3). 

## Statistical analysis 

For each of the three resilience indices studied, robust two-way ANOVAs were used to test for differences between drought events (2005 and 2012) and the two studied oak populations (northern and southern aspects). These tests were used because original and log-transformed data did not follow the assumptions of normality and homogeneity of variance [@Wilcox2012]. Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to the mean value in all cases [@Wilcox2012]. When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers [@Field2012; @Wilcox2012]. Post-hoc differences were assessed pairwise using a similar boostrap test. All the robust ANOVA and post-hoc tests were carried out using the `WRS2` [@Mair2017] and `rcompanion` [@Mangiafico2017] R packages. The level of significance was set to 0.05 and adjusted for multiple comparisons. All analyses were carried out in R software [@R2017] using several packages (see Appendix S7).  


# Results

## Time trends in vegetation greenness

The analysis of greenness time trends showed that EVI in 78.9 % of the pixels of *Quercus pyrenaica* forests experienced a positive trend for the 2000-2016 period. During the 2005 the lowest values of EVI standardized anomalies were recorded, particularly for northern populations (Figure 2). In fact, a "browning" episode ($\mathrm{EVI_{sa}} < -1$) was observed for *Q. pyrenaica* forests during this drought event, yet no changes in greenness were observed in response to the 2012 drought (Figure 3). 

```{r echo=FALSE, results='asis', eval=show_text}
cat("
The analysis of greenness time trends showed that 78.9 % of the pixels of *Quercus pyrenaica* forests experienced an EVI positive trend for the 2000-2016 period, of which 31.67 % were significant trends. The strongest trends were observed in southwestern populations (Figure S2), in which the Cañar site are included Vegetation greenness of *Q. pyrenaica* forests were lower during both the 2005 and 2012 years than the greenness observed for the reference period (Figure 2a). The lowest values for EVI standardized anomalies were recorded in 2005 being singnificantly lower (-2.285 ± 0.029) than those in 2012 (-0.418 ± 0.029) (LSMEANS, t.ratio = -45.358; p_value < 0.0001), particularly for northern populations (Figure 2b). Reduction in annual EVI mean was considerably higher in northern populations than in southern ones during the 2005 drought (Figure 2b). 

According to the standardized anomalies, *Q. pyrenaica* forests suffered a browning episode during 2005 drought event (99.36 % and 79.37 % of the pixels for northern and southern populations respectively), yet no changes in greenness were observed in response to the 2012 drought (Figure S3). 
")
```

```{r echo=FALSE, results='asis', eval=show_text}
cat("
> Cuando exploramos las anomalías (brutas, estandarizadas y normalizadas) observamos valores muy negativos para el año 2005. Sin embargo vemos valores menos negativos para 2005. Tukey posthoc testing (lsmeans package CITAR) was conducted for pairwise comparisons among the slopes and the disturbance years 

> Las anomalías (sa) fueron significativamente menores en 2005 (-2.285 masmenos 0.029) que para 2012 (-0.418 masmenos 0.029), (LSMEANS, t.ratio = -45.358; p_value < 0.0001)
")
```

```{r echo=FALSE, results='asis', eval=show_text}
cat("
## Greenness resilience to drought events

*Q. pyrenaica* forest showed significantly lower resistance (*Rt*) to the 2005 drought event than to that in 2012 (Table 3; Figure 3). Southern populations showed significantly higher values of resistance to drought than northern ones (Table 3), except for 2012 where non-significant differences were recorded (Table S1; Figure 3). 

Recovery (*Rc*) of greenness was significnatly different bewteen drought events and sites (Tables 3). In the 3-year period after the 2005 drought, greenness achieved was 112 % (Rc = 1.12) and after 2012 was 105.7 % (Rc = 1.057) (Table S1). For southern populations, a similar recovery after the 2005 and 2012 drought event was observed (p = 0.2453; Figure 3; Table S1). 

Resilience (*Rs*) was significantly higher for the 2012 drought event than for 2005 (Tables S1, 3), although both values were close to 1 indicating that greenness level was rather similar after each disturbance event (Table S1). The southern populations showed higher resilience values than the northern ones, although these differences were not significant for the 2005 drought event (p = 0.036; Figure 3). 

> Resistencia
> 
> * Los robledales mostraron menor resistencia a la sequía de 2005 que a la de 2012 [2005: 0.858 (0.853-0.863); 2012: 0.943 (0.939 - 0.947); p <0.0001]
> * Menor resistencia de las poblaciones del Norte a los eventos de sequía que las del Sur [N: 0.883 (0.877-0.889); S: 0.921 (0.918 - 0.925); p <0.0001]
> * La resistencia varió en función de la sequía y de la población. Las poblaciones mostraron una resistencia similar al evento de sequía de 2012 (padj = 0.172), sin embargo las poblaciones del N fueron mucho menos resistentes que las del Sur durante la sequía de 2005  [N: 0.819 (0.814-0.824); S: 0.902 (0.896 - 0.907); p <0.0001]
> * notas:  La sequía de 2005 redujo el EVI medio hasta el 85.8%, mientras que la de 2012 la redujo hasta el 94.3%. La sequía de 2005 redujo el evi medio hasta el 81.9 % en las poblaciones del N, y hasta el 90.2 % en las del S. ... drought reduced growth to xx% of the preceding reference period (ver Pretzsch et al 2013 Plant Biology)

> Recovery 
> 
> * La recuperación de los robledales fue mayor tras la sequía de 2005 que tras la de 2012 [2005: 1.120 (1.113-1.126); 2012: 1.057 (1.054 - 1.060); p <0.0001]
> * Los robledales de la cara sur mostraron una menor recuperación que los de la cara norte [N: 1.102 (1.096-1.108); S: 1.069 (1.065 - 1.073); p <0.0001]
> * Las poblaciones del sur mostraron una recuperación similar ante la sequía de 2005 y 2012 (p = 0.186), cosa que no ocurrió para las poblaciones N (p < 0.0001), que mostró una recuperación mayor para la sequía de 2005 que para la de 2012 [2005: 1.169 (1.161-1.177); 2012: 1.042 (1.036 - 1.047); p <0.0001]. En 2005, las poblaciones del S mostraron menor recuperación; mientras que en 2012 ocurrión un patrón inverso, mostrando un patron mayor que las del norte. 

> * La resiliencia de los robledales fue mayor para la sequía de 2012 que para la de 2005 [2005: 0.958 (0.955-0.962); 2012: 0.995 (0.991 - 0.998); p <0.0001]  
> * Los robledales del sur mostraron mayor resiliencia que los del norte [N: 0.970 (0.966-0.974); S: 0.983 (0.980 - 0.986); p <0.0001], aunque para 2005 ambas poblaciones no mostraron diferencias en la resiliencia (padj = 0.152). En 2012 se observó mayor resiliencia en las del S que en la del N (p<0.0001)

> Relative Resilience (De esto no he dicho nada)

  * Los robledales mostraron mayor resiliencia relativa a la sequía de 2005 que a la de 2012 [2005: 0.099 (0.095-0.105); 2012: 0.053 (0.050 - 0.056); p <0.0001]  
  * Las poblaciones del sur mostraron menor resiliencia relativa que las del norte [N: 0.086 (0.082-0.092); S: 0.063 (0.060 - 0.066); p <0.0001], debido sobre todo a la diferencia en la resiliencia relativa para las poblaciones del norte entre los dos eventos de sequía (mucha mayor resiliencia relativa en 2012 que en 2005 para las poblaciones del N)
  * Las poblaciones del sur no mostraron diferencias en cuanto a la resiliencia relativa entre los dos eventos de sequía (padj = 0.152)
")
```

## Radial growth trends and growth disturbances 

The trees of the southern population were older than northern ones, particularly for the high elevation site (CA-High) which had bigger and taller trees than the other sites (Tables 1, 2). Competition was similar among sites but plot basal area was greatest in CA-High (Table 1), which however also exhibited the highest individual BAI growth (Figure 4). Trees for southern sites showed significant positive growth trend since the late 1970s (Figure 4), being more pronunced for the highest site (CA-High). We also observed differences in growth for southern sites. CA-Low growth was lower than that of CA-High but similar to growth of the northern population (SJ) (Figures 4 and S6). 

Las sequías produjeron reducciones de crecimiento en los tres sitios. De los eventos de sequía analizados, destaca por un lado que 1995 produjo la mayor reducción de crecimiento. 
VAS POR AQUI 
For 2005 and 2012 we found a greater reduction of RWI for northern site (SJ) but weaker for southern sites (CA-High and CA-Low). Yet, the lowest pointer year (*i.e.* the lowest growth) since 1950 was 1995 in all chronologies (Figure S3). 

```{r echo=FALSE, results='asis', eval=show_text}
cat("
Sorprende que en los sitios del sur (caH y caL), en los últimos años ha habido crecimientos muy grandes. Por ejemplo, en 2010 se observó uno de los mayores valores de RWI de toda la cronología tanto en caH como en caL. Algo similar ocurrió en sj para los años 2003 y 2013. 
")
```

The analysis of growth changes revealed differences in forest history between sites (Figures 5 and S4). Northern site (SJ) showed two release events (GC > 50 %) detected at stand-wise scale (occurring in more than 50 % of sampled trees): the first during the 1940 decade and the second in the period 1995-2000. These periods alternate with periods of supression. Southern sites (CA-High and CA-Low) showed no release events except for CA-High at the beginning of the 1830 and no suppression event in the last 50 years.

```{r echo=FALSE, results='asis', eval=show_text}
cat("

En las poblaciones del norte osbervamos dos eventos de PGC por encima del 25 e incluso el 50 %. El primero de ellos desde 1940 hasta 1950, mientras que el segundo entre 1994 y 2001. Estos periodos se alternan con periodos de NGC: previo 1940, 1960 - 1991, y quizá aunque no lo vemos en al actualidad (?? duda. Esto es especular un poco). Esto responde creo que a: 

(i) actividad minera y tratamientos forestales. La actividad minera en la zona fue intensa (aunque a veces intermitente, entre 1890 y 1957 (cuando se clausuró); y previamente desde 1858 tras los trabajos de Amalio Maestre). Hay costancia de la existencia de hornos de fundición de mineral en alguna de las minas (La Probadora).
(ii) actividades forestales:
* Se conoce la existencia de un proyecto redactado en 1943 de repoblación forestal de la cabecera del río Genil (he pedido los mapas).
* Ver actividades forestales (bd proyectos de gestión)

En el trabajo de [@GeaIzquierdo2014] se observa una liberación posterior a 1950 (QPUY9) (entre 1950 y 1960 aprox.) y luego una supresión hasta mediados de 1990. Esto coincide con lo que hemos observado en nuestra chrono. 

Respecto a las poblaciones del sur, destaca en la caH una primera liberación en torno a 1829-1830 y 1837-1840; y luego predominan los eventos de supresión que se alternan con algunos de liberación (muy débiles) hastas 1950. Posteriormente se alternan liberaciones-supresiones débiles. Destaca una liberación en torno a 2000-2003 (relación con clima??? posterior a sequía 1999???)

En caL, observamos un periodod de limitación hasta 1885 aprox. Luego se alternan al igual que antes periodos de liberación con otos de supresión con señales más fuertes que en caL, y al igual que antes destaca con mas intensidad si cabe, una liberación entre 2000 y 2003 (antrópico o climática ??)

Duda: ocurre en los dos sitios caL y caH, entonces, es aumento de crecimiento debido a actuación forestal? o debido a clima?? 

[@GeaIzquierdo2014] encontró un pico de disturbance en torno a 1860 en QUPY10 (cañar) sugiriendo que la baja densidad en esos sitios fue presumible el resultado de un thinning (aclareo). El patrón de supresión que observa para este sitio va en la linea de lo que nosotros encontramos (supresiones que alcanzan a veces el 25 % y otras no)


En cuanto a la señal a nivel de sitio, para el Norte, vemos que tanto las supresiones como las liberaciones se observan para mas del 50 % de los árboles. En el caso del sur, vemos que algunas supresiones se observan para mas del 50 % de los árboles, mientras que las liberaciones observadas en torno a los 2000 se observan para mas del 50 % de los árboles 
")
```

## Resilience of greenness and radial tree-growth to drought events
At long-term perspective, the strength of the response of the resilience indices was related to drought severity. The exploration of resilience metrics of tree-growth for all severe droughts in our climatic data (1900 - 2017) (Table S3), revealed a positive relation between resilience (*Rs*) and drought severity ($r^2$ = 0.336, p = 0.048) (Figure 6). A similar pattern was found for recovery (*Rc*; $r^2$ = 0.438, p = 0.019) but negative for resistance (*Rt*, $r^2$ = 0.155, p = 0.205). 

For greenness and tree-growth, resilience metrics were significantly different bewteen drought events at short-term (Table 3). We obtained significantly higher Resilience (*Rs*) and Resistance (*Rt*) values for the 2012, the most severe event, than for 2005 in both variables (Tables S1-S2; Figure 7). Recovery (*Rc*) showed contrasting pattern for EVI and tree-growth. Recovery values for greenness was higher for 2005 than for 2012 drought event (Table S1, Figure 7). 

Resilience metrics significantly varied between sites, except for Resilience (*Rs*) of tree-growth (p = 0.534; Table 3). Southern populations showed lower values for Recovery (*Rc*) than northern site for greenness and tree-growth (Tables S1 and S2). In opposite, Resistance and Resilience values were higher for southern sites than for northern ones (Tables S1 and S2).













 













```{r echo=FALSE, results='asis', eval=show_text}
cat("
> Resistance: 

Although no significant differences were observed in the resistance (*Rt*) of oak radial growth between the two drought events (Table 3), the 2005 drought reduced growth more than that of 2012 (Rt = 0.721 and 0.819 respectively) (Table S2). Similar to results for greenness, the northern site, which is under a drier climate, showed resistance values lower than those of the southern site, especially for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figure 3). 
  
* No diferencias significativas entre la resistencia mostrada por los robledales a los eventos de sequía de 2005 y 2012 (F = 6.0189; p = 0.019). Valores de resistencia menor al evento de 2005. [2005: 0.721 (0.6437-0.7984); 2012: 0.8193 (0.7758 - 0.8628); p = 0.025; trimmed-p = 0.019]

* Diferentes resistencias a los eventos de sequía en función del sitio (p <0.0001). Las localidades del sur (CaLow y CAHigh no mostraron diferencias en cuanto a la resistencia p.adjust = 0.012) mostraron una resistencia mayor a los eventos de sequía que la observada a la localidad del norte [caH: 0.8157 (0.7549 - 0.8764) (a); caL: 0.9209 (0.8834 - 0.9584) (a); SJ: 0.6116 (0.5387 - 0.6846) (b)].

* La interacción también fue significativa. De hecho, si miramos las gráfica, observamos como la resistencia a la sequía de 2005 fue significativamente menor en SJ que la resistancia mostrada, tanto por las otras poblaciones para 2005, como la mostrada por SJ para el año 2012 --¿¿¿ Podemos decir que la sequía de 2005 afectó mucho mas a la población de SJ??? 

> Recovery 
> * Mayor recuperación del crecimiento de los robledales tras la sequía de 2005 que tras la de 2012 [2005: 0.9462 (0.8794-1.013); 2012: 1.161 (1.081 - 1.24); p < 0.001]
> * Las poblacionees del sur (caH y caL; no diferencias entre ellas) mostraron una menor recuperación que las poblaciones del norte, es decir, las del norte se recuperaron mas rápidamente que las del sur que no llegaron incluso a valores de Rc = 1. 

Similar pattern of resilience (*Rs*) values was found for growth than for greenness respect to drought event: significantly higher values of *Rs* for the 2012 drought event than for the 2005 (Table S2; Figure 3). However, no differences were observed between sites (Table 3). For 2005 drought event, *Rs* value of SJ (northern site) was lower than that of southern ones (CA-High and CA-Low), but opposite pattern was found for the 2012 drought event. All valu`es of *Rs* for growth were below 1, except for the SJ site in 2012 (Rs = 1.031). 




YA ESTOY DESESPERADO CON ESTE PAPER
*Q. pyrenaica* forest showed significantly lower resistance values (*Rt*) to the 2005 drought event than to that in 2012 for greenness and for radial growth (Table 3; Figure 7). The 2005 drought reduced greenees and growth more than that of 2012 (Tables S1 and S2). Resistance values to drought for greenness and tree-growth varied between sites (Table 3). Southern populations showed significantly higher values of resistance than northern ones (Tables S1 and S2). It was particularly important for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figure 7). 

Recovery (*Rc*) of greenness and growth were significantly different bewteen drought events and sites (Table 3). In the 3-year period after the 2005 drought, greenness was 112 % (Rc = 1.12) and after 2012 was 105.7 % (Rc = 1.057) (Table S1). An opposite pattern was found for tree-growth, with significantly lower values of recovery after the 2005 drought, staying at levels of *Rc* < 1 (Figure 7, Table S2).


Northern populations showed significantly higher values of recovery than southern sites for greenness and tree-growth (Table S2). For southern populations, no significant differences were found for recovery of greenness after the 2005 and 2012 drought event (p = 0.2453; Figure 7; Table S1). Recovery values for tree-growth of southern populations were below or close to 1 (Figure 7, Table S2).   

Significantly higher values of resilience (*Rs*) were observed for the 2012 drought event than for the 2005 in both variables (greenness and tree-growth) (Tables S1-S2; Figure 7). Resilience values varied significantly between sites for greenness, but not for tree-growth (Table 3). Southern populations showed higher resilience values (*Rs*) than the northern ones (Tables S1-S2). For greenness, the differences of resilience between sites were not significant for the 2005 drought event (p = 0.036; Figure 7). For tree-growth, opposites resilience values were found for the interaction between sites and drought event: higher values of resilience for northern populations than southern ones (CA-High and CA-Low) during the 2012 drought event but opposite pattern during the 2005 (Table S2). 



Similar to the long-term observed pattern, 

was greater after more severe drought event (2012). Opposite pattern was observed for greenness with 

Lower resistance values (*Rt*) were observed to the 2005 drought event than to that in 2012 for both variables. The 2005 drought reduced greenees and growth more than that of 2012 (Tables S1 and S2). Resistance values to drought for greenness and tree-growth varied between sites (Table 3). Southern populations showed significantly higher values of resistance than northern ones (Tables S1 and S2). It was particularly important for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figure 7). 


") 
```

## Climate and tree-growth
Tree-growth exhibited a significant greater response to water availability than to temperatures, as it can be expected for a Mediterranean site and a species located closer to its xeric/dry distribution limit (rear-edge). Cumulative precipitation of the hydrological year was the climatic variable with the highest relationship with growth and influenced similarly and positively the tree-growth for both northern and southern populations (Figure 8a). Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population. Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 8a), specially for the northern population (r > 0.6). Relationships with temperatures were lower than those with moisture variables. Yet there were differences between northern and southern populations: spring maximum temperature was the most significant limitation for the tree-growth of the southern population at both elevations while no significant limitations were recorded for northen site regarding the temperatures (Figure 8b and 8c). Remarkably minimum and maximum temperatures of current September positively influenced the tree-growth only in the northern population. No positive relationship with temperature was found for southern trees. 


# Discussion
## Rear-edge oaks show high resilience in response both to recent drought events and long-term climatic variability. 

Our findings show that severe droughts, such us 2005 and 2012, provoked a reduction both in greenness (*i.e.* primary growth) and in secondary growth of *Q. pyrenaica* (Figures 2 and 4). These results are consistent with previous studies that reported significant reductions of tree-growth for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014; @RubioCuadrado2018]. The greatest reduction of tree-growth was observed during the 1995 drought, a characteristic pointer-year that caused severe and extensive damage to the Mediterranean vegetation across Iberian Peninsula [@Penuelas2001; @Camarero2018; @Gazol2018]. 

During the last decades, the south of Iberian Peninsula suffered two of the worst recorded drought [2005 and 2012; @GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3). However, we found a positive trend for vegetation greenness of *Q. pyrenaica* (78.9 % of the pixels showed positive trend  during the 2000 to 2016 period). This confirms previous findings that pointed out an increase in primary productivity for these forests [@PerezLuque2015onto], being the most productive ecosystems in this mountain region during that period [@Alcaraz2016obsnev_ndvi]. For tree-growth, we also observed positive trends in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4). Similar long-term trends were described for this species along their distribution range [@GeaIzquierdo2014; @RubioCuadrado2018] and could be related to the rising of the temperatures in the last decades [@GeaIzquierdo2014]. Our pattern differs from the decline trends observed for several mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P. sylvestris*; @SanchezSalguero2012; @Camarero2015; *Fagus sylvatica*; @Dorado2017].

Resilience (*Rs*) values observed in our study sites for the latest drought events in both tree-growth and greenness were close to 1, except for the tree-growth during 2005 event (Tables S1 and S2). Despite the 2012 drought event was more severe and intense than 2005, according with our analysis with SPEI data (Table S3), resilience values for greenness and tree-growth were greater for 2012 than for 2005. These results could be explained by the different timing of the two droughts, *i.e.* the 2012 drought was a winter drought [@Trigo2013] that occurred earlier than the 2005 drought. The latter lasted less than 2012 drought, but matched the period of maximum growth for *Q. pyrenaica* forests [@PeresLis2017]. Our results are in line with recently studies that indicate the timing of the drought as a key factor determining tree recovery after drought [@Huang2018],  that also has been observed for other Mediterranean oak species [*e.g.* Holm oak; @Camarero2015b]. 

As we mentioned previously, we found strong declines of tree-growth during the most severe drought events occurred (1995 and 1999) (Table S3; Figure S3). This pattern was also observed for other species in their rear edge [@SanchezSalguero2012; @Camarero2015]. But, interestingly, we obtained the highest values of resilience for tree-growth for these two drought events (Figure 6). In addition, the analysis from a long-term perspective of the severe droughts suggests a positive relation between the tree-growth resilience and drought severity (Figures 6). Surprisingly this result seems contradict our initial hypothesis in which we predict low resilience values for this oak species in their rear-edge after extreme drought events. 

Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of the species [@Hampe2005]. For instance, higher vulnerability to drought is ofted assumed for rear-edge populations [@MartinezVilalta2018], however some studies showed different findings [@Cavin2017; @Granda2017], as our results suggest, with high values of resilience for populations of * Q. pyrenaica * located in its rear-edge.

This would be related with the definition of the marginal population. Algunos autores han apuntando que cuando se estudian poblaciones del rear-edge, hay que poner atención a la forma en la que se define la marginalidad, esto es, si se define atendiendo a criterios geográficos, climáticos, o según otros factores ecológicos [@MartinezVilalta2018]. En este sentido, los altos valores de resiliencia a los eventos de sequía que hemos observado, podrían sugerir que las poblaciones de *Q. pyrenaica* en Sierra Nevada están situados en un rear-edge geográfico, pero no climático. Esto último se ve reforzado por el hecho de que Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010]; además las poblaciones de *Q. pyrenaica* en Sierra Nevada presentan una alta resiliencia genética [@Valbuena2013; @Valbuena2017]







Several considerations have to be taken into account to explain this result. Firstly it would be interesting to explore the cumulative precipitation of the previous and posterior years of these drought events. For instance, the 1992-1994 years (pre-1995 drought) had precipitation values below the mean of the whole period (Figure S1), while 1996-1998 years (post-1995 drought) showed a wet opposite pattern. In fact, the first months of 1996 recorded the highest cumulative rainfall of the past century for our southern site (Cáñar) [@MesaTorres2009]. In addition, this argument could also explain the higher values of recovery found for 1995 (Figure 6). 

A plausible explanation could be 

--> Ver y resumir páginas 18-19 Notas Naturkunde 


A remarkably result is the higher values of resilience for tree-growth obtained during the worst and more severe drought events in our study area (1995 and 1999) (Table S3; Figure 6). A plausible explanation could be the water availability after the drought event. For instance, the months with the highest cumulative rainfall of the past century occured during the first months of 1996 in our southern site (Cáñar) [Torres-MESA]; and years after 1995 were within the wetter of the past 60 years. 



--> Aquí iría la idea de lo de refugio 

---> Nota 

Para la discussion: 
- Resilience values of RW for 2005 was the lowest of the drought events analized even not being the moste severe drought event 
- 2005 reduced greenness and bai more than 2012 

```{r echo=FALSE, results='asis', eval=show_text}
cat("
Our findings show that recent severe drought events, such as 2005 and 2012, provoked a reduction both in greenness (*i.e.* primary growth) and in secondary growth of *Q. pyrenaica*. 


During the 2005 drought, one of the worst drought events recorded in the Iberian Peninsula [@GarciaHerrera2007], we found a browning of the *Q. pyrenaica* forests, but no changes in EVI standardized anomalies were recorded for 2012, which can be explained because 2012 drought event was a winter-drought [@Trigo2013]. 


Tree-growth was also affected by drought as evidenced by the reduction in detrended tree-rings (RWI) during the most severe drought events (Figure S3). The decline in growth observed in our study sites is consistent with several works reporting tree-growth reductions for Mediterranean tree species during severe droughts, particularly for 2005 [*e.g.* @SanchezSalguero2013; @Camarero2018; @Gazol2018]. 

Although 2005 and 2012 were two severe droughts recorded for the south of the Iberian Peninsula [@GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of the forests of *Q. pyrenaica* located in their rear edge. Our results agree with those obtained by previous works using other remote-sensing vegetation indices [@PerezLuque2015onto; @Alcaraz2016obsnev_ndvi], which suggests an increase in primary productivity during the last years for rear-edge populations of this species. 


For tree-growth, a similar positive trend was observed in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4). This result differs from those previous reported for *Q. pyrenaica* along their distribution range [@GeaIzquierdo2013; @GeaIzquierdo2014]. @GeaIzquierdo2014 found a general decline in the growth of this species since the 1970s, particularly sharp for populations located in their dry-edge. This decline trend in growth have also been oberved for other tree species located in their rear-edges [*e.g.* @SanchezSalguero2012; @SanchezSalguero2017; @Dorado2017]. Furthermore, growth projections have forecasted a decrease in productivity for *Q. pyrenaica* that would increase vulnerability of this species to climate warming at the dry edges locations [@GeaIzquierdo2013]. However, our results for similar locations have revealed a recovery in tree-growth for the last years (Figure 4). In addition, a similar positive trend in BAI for the last years, has been reported for another *Q. pyrenaica* population of the Sierra Nevada [@RubioCuadrado2018]. 

Aunque los valores de resiliencia (*Rs*) para el greenness y el tree-growth fueron inferiores o muy cercanos a 1, observamos en ambas variables, una mayor resiliencia tras la sequía de 2012 (Figure 7). La sequía de 2012, que fue mas severa e intensa que la registrada en 2005 (Table S3), ocurrió durante el invierno de 2012 [@Trigo2013], lo cual puede explicar un menor efecto, sobre todo en el greenness de *Q. pyrenaica*. Por otro lado, al analizar la resiliencia del tree-growth para otros eventos de sequía (long-term perspective), observamos como estas poblaciones presentan altos valores de resiliencia. Como muestran nuestros resultados, los mayores valores de resiliencia fueron registrados para dos de los eventos mas severos de sequía (1995 y 1999), con valores de resiliencia por encima de 1.2 en ambos eventos (Figure 6). 
") 
```

Desde una perspectiva a largo plazo, es destacable el hecho de que las poblaciones de *Q. pyrenaica* en Sierra Nevada, aún habiendo sufrido varios eventos de sequía severa, presentan altos valores de Recovery y de Resiliencia, tal y como indican nuestros resultados (Figure 6). 


```{r echo=FALSE, results='asis', eval=show_text}
cat("
Hemos observado como las sequías severas, sobre todo la de 2005, provoca una reducción en el greenness y pero sobre todo en el crecimiento de *Q. pyrenaica*. Los datos de anomalías estandarizadas de EVI mostraron un browning para la mayoría de los robledales de Sierra Nevada durante la sequía de 2005. 

La sequía de 2005 ha sido una de las mas severas de los últimos años, afectando significativamente al crecimiento

#--------------------------------------------
# Sequías 

### 2005 frente a 2012 

* Analizando los datos de precipitación acumulada de las estaciones alrededor de SN, observamos como 2005 fue el año mas seco desde 1950. 
* Datos de SPEI (0.5 º Grid) 
    * 2005 fue peor que 2012, pero no peor que 1995 o que 1999 (ver figura S2)
    * Tabla S4. Severidad sequías, la de 2005 fue menos severa y menos intensa y de menor duración que la de 2005 (de donde vienen estos datos???)

* La sequía de 2005 fue una de las mas severas afectando significativamente al crecimiento. Esto se ha observado también en otras especies en el sur de la P. Ibérica (p.ej. P. nigra en Andalusia [@SanchezSalguero2013], ... INCLUIR otras especies y citas).


### 2005 en contexto 
  
* Algunos trabajos hablan que los peores años de sequía en la P. Iberica durante la segunda mitad del siglo XX fueron 1986, 1994-1995, 1999, 2005 y 2012

* An analysis of the long term series from meteorological stations (n=54) of Iberian Peninsula (1961-2011) reveals that major drought episodes in the Iberian Peninsula were recorded in 1981, 1995, 2000 and 2005 [@VicenteSerrano2014].


### Reducción en el crecimiento de algunas especies 
El efecto de la sequía de 2005 en el crecimiento también se ha visto en otras especies en el sur de la P. Ibérica 
La sequía de 2005 fue una de las mas severas afectando significativamente al crecimiento. Esto se ha observado también en otras especies en el sur de la P. Ibérica (p.ej. P. nigra en Andalusia [@SanchezSalguero2013], ... INCLUIR otras especies y citas).

* P. nigra: Se observó una gran reducción de crecimiento en 1994-1995, 1999 y 2005 para Pinus nigra en el SE de Andalusia [@SanchezSalguero2013] coincidiendo con años de sequía para esa zona. La máxima reducción en crecimiento fue observada para 2005, siendo significativamente mayor que para el resto de años [@SanchezSalguero2013].


No obstante, a pesar de que 2005 y 2012 fueron dos de las sequías mas severas registradas para el sur de la P. Ibérica [@GarciaHerrera2007; @VicenteSerrano2014] (Appendix S3), nuestros resultados muestran que los robledales de *Q. pyrenaica* en su rear-edge presentan una tendencia positiva tanto en el greenness (78.9 % pixeles con greeness) como en el crecimiento (tendencias positivas del BAI). 


### La resiliencia de las poblaciones de roble de Sierra Nevada. 

El greennness de los robledades en Sierra Nevada ha mostrado una tendencia positiva hacia mas verdor en los ultimos años, que coincide con lo que ya observamos con datos de NDVI [@PerezLuque2015onto], sobre todo para las poblaciones del sur. Esta tendencia positiva, obtenida de variables derivadas de remote sensing (EVI y NDVI), parece que también se observa en el crecimiento. Por ejemplo en las poblaciones del sur (CaHigh y CaLow) observamos en los últimos años una ligera tendencia de crecimiento. 


Aún habiendo pasado varios periodos de sequía severa (sobre todo 2005, el crecimiento -BAI- en el N, se redujo hasta el 45 %), estos robledales han mostrado una alta resiliencia. Por ejemplo para las poblaciones del norte, los valores de EVI durante la sequía de 2005 descendieron hasta el 81 %, mientras que el BAI (sitio SJ) lo hizo hasta el 45 %. Sin embargo, la recuperación fue rápida, así los valores de Recovery (Rc) para el EVI tras la sequía de 2005 (en las poblaciones del norte) fue de 1.17, mientras que para el BAI el Rc fue de 1.112. En definitiva, estamos observando altos valores de resiliencia en estos robledales.


Despite the resilience values (*Rs*) of growth for 2005 and 2012 were close or below 1, we observed high resilience of these forests from a long-term perspective. As our results indicate, the highest resilience values for growth were recorded for two of the most severe drought events (1995 and 1999), with resilience values above 1.2 in both events (Appendix S5).

genetica 

Los robledales de SN estén situadas en su borde rear-edge donde se supone que sufren mas estrés climático. Hemos observado un aumento en el greenness en los últimos años. Por otro lado, hemos encontrado que son poblaciones resilientes a la sequía, tanto para el crecimiento como para el greenness. Además, estos robledales tienen una alta resiliencia genética  ¿Sierra Nevada (regiones de montaña) como refugio?? quizá este rear-edge esté actuando como refugio?? (esto es muy especulativo)


Las zonas rear edge son zonas que se supone son mas vulnerables al cambio climático por sus estrechos márgenes. De hecho como han visto algunos trabajos, las especies situadas en su rear edge son mas vulnerables. No obstante tambiéh habría que considerar que algunos trabajos apuntan a que la vulnerabilidad dentro del rear edge es mas sitio dependiente (dorado). Además habría que introducir aqui lo que M. Villalta habla de la consideración de rear edge de.


* Otro punto de interés a introducir aquí es que frecuentemente se asume una alta vulnerabilidad a la sequía de las poblaciones situadas en su dry rear-edge [@MartinezVilalta2018], sin embargo algunos estudios están demostrando esto no es siempre así (ver por ejemplo [@Cavin2017; @Granda2017]). Nuestros resultados creo que van en esta línea creo. Esto, como apunta [@MartinezVilalta2018]m tiene que ver con la que consideramos como habitat marginal de la especie (... When the focus is on marginal, rear-edge populations, proper consideration should be given to the different ways in which marginality can be defined (stressing geographical, climatic or other ecological factors). 

¿Casos parecidos?

Existen trabajos que han analizado los efectos de la sequía en especies situadas en su rear edge (Asier, por ejemplo; Matias) y algunos de ellos se han centrado en ver como responde el crecimiento de las especies en el rear edge utilizando dendro (Herrero Rigling.. // dorado-liñan // sanchez salguero  

trabajos rear edge: 
 * sanchez salguero Is drought the main decline factor at the rear edge of Europe? The case of southern Iberian pine plantations 
 * sanchez salguero Assessing forest vulnerability to climate warming using a process-based model of tree growth: bad prospects for rear-edges
 * dorado-liñán: Climate threats on growth of rear-edge European beech peripheral populations in Spain.
 * herrero rigling: Varying climate sensitivity at the dry distribution edge of Pinus sylvestris and P. nigra
 * dorado-liñan: Large-scale atmospheric circulation enhances the Mediterranean East-West tree growth contrast at rear-edge deciduous forests. 
buscar otros 

... Varios trabajos han apuntado la existencia de vulnerabilidad local al aumento de temperaturas en los sitios del sur de distribución de esta especie [@GeaIzquierdo2013; @GeaIzquierdo2014] así como una strong response to moisture availability...; Además teniendo en cuenta que en los últimos años estamos viendo un aumento en las sequías (menor disponibilidad de agua) --> por ello queremos analizar como están respondiendo esta especie, sobre todo en esas zonas dry-egde, y analizar su resiliencia... 

## refugio?? 

Los robledales de SN estén situadas en su borde rear-edge donde se supone que sufren mas estrés climático. Hemos observado un aumento en el greenness en los últimos años. Por otro lado, hemos encontrado que son poblaciones resilientes a la sequía, tanto para el crecimiento como para el greenness. Además, estos robledales tienen una alta resiliencia genética [@Valbuena2013; @Valbuena2017]. ¿Sierra Nevada (regiones de montaña) como refugio?? quizá este rear-edge esté actuando como refugio?? (esto es muy especulativo)

pero ojo, quizá el problema es que no estamos definiendo bien el rear edge --> ver martinez vilalta et al.  




Las poblaciones situadas en su rear edge viven en estrechos márgenes ambientales y pequeños variaciones en las condiciones ambientales pueden provocar que las restricciones ambientales sean mas severas [@Hampe2005]. Así, frecuentemente se asume una alta vulnerabilidad a la sequía de las poblaciones situadas en su dry rear-edge [@MartinezVilalta2018], tal y como se ha visto para *Q.pyrenaica* [@GeaIzquierdo2014] y para otras especies (citas). Sin embargo, algunos estudios están demostrando que esto no es siempre así [ver por ejemplo @Cavin2017; @Granda2017], tal y como sugieren nuestros resultados, con altos valores de resiliencia para poblaciones de *Q. pyrenaica* situadas en su rear-edge. 

Algunos autores han apuntando que cuando se estudian poblaciones del rear-edge, hay que poner atención a la forma en la que se define la marginalidad, esto es, si se define atendiendo a criterios geográficos, climáticos, o según otros factores ecológicos [@MartinezVilalta2018]. En este sentido, los altos valores de resiliencia a los eventos de sequía que hemos observado, podrían sugerir que las poblaciones de *Q. pyrenaica* en Sierra Nevada están situados en un rear-edge geográfico, pero no climático. Esto último se ve reforzado por el hecho de que Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010]; además las poblaciones de *Q. pyrenaica* en Sierra Nevada presentan una alta resiliencia genética [@Valbuena2013; @Valbuena2017].

")
```


## Site environment shapes differential sensitivity to climate and drought of rear-edge oak populations 

Our results showed differences for greenness and tree-growth between northern and southern oak populations (Table 3). Las poblaciones del norte, que presentan condiciones mas secas (Table 1), se vieron mas afectadas por los eventos de sequía. For instance, the northern site showed more negative EVI standardized anomalies (higher browning intensity) than the southern sites during the 2005 drought event (Figure 2). In addition, the stronger correlations of tree-growth with SPEI (Hydrological and summer) observed for northern site (Figure 8), can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014]. 

Moisture availability has been reported as the most limiting factor driving radial growth of *Q. pyrenaica* along their distribution range in Iberian Peninsula [@GeaIzquierdo2014]. We found a singnificant positive relation between precipitacion (hydrological year and previous December) and tree-growth (Figure 8a). Our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth [@Roig2009; @GeaIzquierdo2014; @GeaIzquierdo2014FEM; @Leal2015; @Camison2016; @GonzalezGonzalez2014; @GarciaGonzalez2017]

A striking result is the difference for tree growth between sites (Figure 4). The trees of CA-High site, which are located around 1900 m. *a.s.l.* and representing the upper limit of the treeline of the species in this southernmost location [@PerezLuque2011tfm], showed higher BAI than those located at low-elevations sites (CA-Low and SJ) (Figure 4). This is especially interesting for southern sites, which are very close to each other. Our results are consistent with previous findings that pointed out tree growth and tree responses to drought are site-dependent (e.g. soil features, tree competence, etc.) [e.g. @Babst2013; @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017; @Dorado2017c]. 

When we explored the chracteristics of the southern sites, we firstly observed a difference in elevation. @Bhuyan2017, in a recent work across the European continent, found a clearly effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive. The higher values of tree growth that we observed in CA-High site could be related with lower water stress at high-elevation sites, since moisture availability is a key factor limiting tree growth for *Q. pyrenaica* [@GeaIzquierdo2014]. In fact, the differences in available water of soils could explain our results. @CoboDiaz2017, in a work analyzing the variation of the microbiota along an elevational gradient within this location, found higher values of available water of the forest soils located on high elevation compared with those located at low elevations. In this sense, a remarkable observation is the presence of traditional irrigation ditches which could provide an additional water supply. In most of the high-mountain watersheds of Sierra Nevada, there is a system of historical irrigation channels, know as *acequias de careo*, that was used since Middle Age to cultivated these valleys [@MartinCivantos2014; @MartinCivantos2016]. These ditches run through the hillsides of valleys releasing water through several points that filtrate the water to recharge the aquifer and then irrigates the foot of the slopes [@MartinCivantos2014; @Jodar2017]. The network of traditional ditches present in one of our sampling site (Cáñar) was recentlty described, with detailed information about water use [@RuizRuiz2017] and its hydrological functioning [@MartinMontanes2015]. There is a ditch (*Acequia de la Era Alta*) located uphill the CA-High site, which functions from March to June [@RuizRuiz2017]. It brings water from snowmelt and from a nearby stream, soaking the *Q. pyrenaica* forests. This traditional system could supply an extra of water that could be used by the trees located downstream of these channels. This extra of water is particularly important for trees of this dry-rear edge and could explain the higher values of BAI for trees located in CA-High site. 

## Land-use legacy effects shape distribution and sensitivity to climate change of read-edge oak populations

The Mediterranean mountains, even at high elevations, have suffered strongly transformation of the landscapes driven by human activities throughout history [@Regato2008]. Multiple evidence from palaeoecological studies indicated an intense human impact on vegetation of Sierra Nevada since 3000 cal year BP [@Anderson2011; @JimenezMoreno2013; @GarciaAlix2017; @MesaFernandez2018]. From that moment onwards, increases on fire occurrence and both grazing and mining activities, were recorded for this mountain region, with an sharp intensification of the human activities in the last 150 years [@GarciaAlix2017; @MesaFernandez2018]. Mining, forest clearings, fuelwood and charcoal exploitations, pastoralism and also wars, have strongly impacted on the forest resources in Sierra Nevada. As a consequence of all those activities a loss of about 90% of broadleaf *Quercus* species cover from medium and low elevation occurred in this mountain region [@JimenezOlivencia1991]. 

Our study sites have been subjected to an intense exploitation of forest resources. Although the estimated age for our sampled trees was not more than 180 years (Tables 1, 2), several documents reported the presence of oaks before. For instance, the inventories of trees made by the Spanish Navy during the second half of 18th century, recorded the quantity of trees, dividing them into three categories: *new*, *growing* and *old* [@Wing2015]. For Cáñar site more than two millions of trees were reported, most of them *news*, and no *old* trees were counted, suggesting recent wood fellings. Less quantity were reported for San Juan location (circa 700 000 trees), which 220 *old* trees and 56 700 *growing* [@GomezCruz1992]. These numbers are a reflect of the intense exploitation of the forest resources occurring up to the end of nineteenth century [@ValbuenaCarabana2010; @Calatrava2019] and also concur with the regeneration peaks observed for several Iberian oak woodlands [e.g. @GeaIzquierdo2014; GeaIzquierdo2015; @Dorado2017].  

We found similar tree competence levels in our sampled sites (Table 1), but differences in tree size and age suggest different management origin (*i.e.* land-use legacy). Our study sites had different land-use trajectories driven by the differential pattern of the natural resource use [@JimenezOlivencia2015; @MartinCivantos2014]. 

In Northern sites (e.g. San Juan), there is a sequentially distribution of the land uses along the elevational gradient [CITA CATALOGO]. Grasslands and shrublands for cattle farming are located at high elevations. Then forests formation with some croplands; and then, at lowlands, irrigated terraces with tree crops. Another activities could conditioning the natural resources uses. For instance, the mining activity in San Juan site have strongly conditioned the forest structure. This oak woodland is located in an area with a high concentration of mines and quarries that have been exploited intermittently throughout history (citas).  Historical documents indicated two periods of intense mining activity: the second half of the 19th century after the publication of detailed mineralogical reports by [@MaestreXXXX]; and during the first decades of the twentieth century until 1960, which is the last year in which there is evidence of the existence of mining activity [cita].  

On the other hand, woodland areas of the southern slopes (Cáñar) are mixed with a greater percentage of croplands, even reached high elevation (mainly barley, rye and potatoes) [@Calatrava2019]. Forest resources, like firewood, charcoal, acorns, have been continuously exploited in this site through history [@ValbuenaCarabana2014, @JimenezSerrano2004]. In Cáñar there were an arraigated tradition of charcoal extraction since fifteenth century [@JimenezSerrano2004] up to middle of the 20th century. From this moment to the present, there were a sharp decrease of the wood extraction, mainly due to rural abandonment [@MesaTorres2009; @Bonet2014_conama].

The different patterns of resource uses can help to understand the tree-growth changes recorded in our chronologies (Figure 5). We observed a tree-growth release event around 1940 in the SJ oak woodland (Figures 5, S4) which concurs with one period of mining activity for this area. Several documentary sources indicate an apogee of the mining activity during 1925 to 1957 period, supported by both the improvement of the paths to transport the mineral and by the creation of new railway connections (CITA). During this period there were an increase on the use of timbers for the tunnels of the mines (CITE) and several furnaces that required great quantities of fuelwood to melt the mineral were active in this area (cite). This heavily exploitation of the forest resources could affect to a major part of this oak woodland, since percentage of trees affected by GC > 50 % reaches values above 50 % (Figure S4). Our results also coincide with a concurrent but less intense tree-growth release event reported for a closed oak woodland [@GeaIzquierdo2014]. In addition, paleoecological studies carried out in nearby alpine bogs have recorded increases in the heavy metals concentration since the end of 18th century until mid-20th, which coincides with the maximum activity of the mining in this area [@GarciaAlix2017]. 

The other release event observed for SJ site during 1995-2000 was lower than the occurred at 1940, but affecting more trees (Figures 5, S4). We revised the forest practices carried out in this area in the last 30 years [CITA], but we did not find any clearing or cutting event during 1995 - 2000 period. Two non-mutually exclusive ways could explain the tree-growth release observed. Firstly it could be related with a natural drought-induced mortality event after 1995, as was reported for Mediterranean tree species [*e.g.* @Penuelas2000; @Lloret2004; @Gentilesca2017]. On the other hand, we obtained strong positive correlations of SPEI with tree-growth for this site (Figure 8), which suggests a high sensitivity to water availability of this slightly more xeric site (Table 1) [@GeaIzquierdo2014]. Previous works reported a sharp decline (or non-production) in latewood production during extreme drought events [@Corcuera2006]. Since latewood are less vulnerable to embolism than earlywood vessels [@Corcuera2006], the reduction or even the non-production of former could negatively affect to tree-growth and also enhanced the mortality, particularly for drier sites [@Corcuera2006; @GeaIzquierdo2014].














```{r echo=FALSE, results='asis', eval=show_text}
cat("

#### Disturbance
Estos periodos se relacionan bien con eventos antrópicos: mineria (el primero de ellos) y actividades forestales el segundo (esta frase de la minería y demás la dejamos para la discusión??). 

## historia forestal de ambos sitios 
Incluimos lo que conocemos de la historia forestal de los sitios?? Me explico, tenemos datos de manejo y uso antrópico de las dos zonas que proceden de varias fuentes. En resumen, mas o menos, tenemos: 

* Minería: datos de minería, que afectan sobre todo a la población de SJ (los tenemos localizados temporalmente)
* Actuaciones forestales: 
* Incendios, Carboneo, Ganadería, etc --> Tenemos información menos estructurada sobre estos ámbitos, procedentes de varias fuentes, que tienen una incertidumbre mayor espacial y temporalmente. 
")

```

Climatic drivers are key factors determining the growth of tree species, especially at the rear edge of their distribution. Para  *Q. pyrenaica* se ha visto que moisture availability es el factor limitante que determina el crecimiento en las poblaciones situadas en su rear edge [@GeaIzquierdo2014], pero para otras especies (*i.e.* *Pinus nigra*, *P. sylvestris*) la temperatura tiene mas peso que la disponibilidad de agua  [@Herrero2013; @Matias2017]. Pero es importante considerar además la historia de manejo que han tenido esos bosques a la hora de la forest management [@Penuelas2017; @DoblasMiranda2017], ya que, como observamos en nuestros resultados puede condicionar el crecimiento y la resiliencia de las especies. 

```{r echo=FALSE, results='asis', eval=show_text}
cat("
Podríamos complementarlo con lo que le pasa a otras especies en su borde de distribución: por ejemplo en Baza, Herrero et al. 2013, encontraron para Pinus nigra y sylvestris que la temperatura tenía mas peso que la disponibilidad de agua). O también ver algunos de los trabajos de Camarero et al 2013 para el P. nigra en su borde de distribución u otros similares (el de Sanchez-Salguero et al. 2013, 2015) 


Para *Q. pyrenaica* moisture availability was reported to be the most limiting factor driving radial growth in Iberian Q. pyrenaica populations [@GeaIzquierdo2014] (Prec hidrológica y SPEI) (ver también Gea-Izquierdo et al. 2015 European Journal of Forest Research). Lo que hemos obtenido aqui (analizando solo el rear edge) también van en esa línea. 

")

```
```{r echo=FALSE, results='asis', eval=show_text}
cat("
* Vicente-Serrano: 
    * We found that some forests from cold and humid areas respond to shorter drought time-scales than forests from dry areas, which usually respond to longer time-scales
    * Growth and responses to drought are modulated by site conditions such as soil type, specific functional traits and the intensity of competition among neighbouring trees (Orwig & Abrams, 1997; McDowell et al., 2008; Linares et al., 2010; Pasho et al., 2011).
    * Our findings provide evidence that the patterns of growth response to drought do not follow a general geographical structure and that these patterns are driven by the biogeographical, topographic and climatic conditions of each site, showing that forests located in different continents have the same pattern of response to drought time-scales.


-----------

Despite similar value of competence in our study sites, we obtained higher values of growth for the highest elevation site (CA-High, Figure 4). 

------ 


En un estudio a escala continental, [@Bhuyan2017] encontraron que para la misma especie (estudiaron 850 sites: Fagus sylvatica, Abies alba, Picea abies, Larix decidua, Pinus cembra, P sylvestris, P nigra, Quercus petraea y Q robur) los stands situados a mayores elevaciones fueron menos sensibles a la sequía que los situados a elevaciones inferiores (stands were less drought sensitive at higher elevations compared to lower elevations). 
The effect of elevation was seen clearly in the case of several species where high elevation sites showed greater drought resistance compared to stands at lower elevation in the same climate zone.

")

```
```{r echo=FALSE, results='asis', eval=show_text}
cat("
## Rear-edge 

* Se asume una alta vulnerabilidad de las especies a la sequía en su límite inferior, sin embargo esta vulnerabilidad no necesariamente se mantiene: The results of Granda et al. 2018 are consistent with recent studies showing that the frequently assumed higher vulnerability of dry, rear-edge populations to drought does not necessarily hold (e.g. Cavin and Jump 2017) (from: Martinez-Vilalta 2018)

* Definición de *marginal* populations. [@MartinezVilalta2018] indica que a veces no está claro la definición de marginal populations. When the focus is on marginal, rear-edge populations, proper consideration should be given to the different ways in which marginality can be defined (stressing geographical, climatic or other ecological factors). 

* Quizá estamos asumiendo que es una población marginal, pero sin embargo lo es solamente geográficamente

----- 

No obstante tambiéh habría que considerar que algunos trabajos apuntan a que la vulnerabilidad dentro del rear edge es mas sitio dependiente (dorado).
 
### Diferencias entre sitios

Llama la atención que en el rear-edge de la distribución de la especie, en el sitio mas meridional dentro de este rear edge, y en la parte mas alta (estas parcelas están en el treeline de la especie en SN (en torno a 1900)), es donde encontramos mayores crecimientos. Parece que los robles estuvieran mejor en esta zona. Algunas reflexiones sobre esto: 

* Quizá están creciendo donde les están dejando (menor impacto antrópico en las zonas mas altas, al menos en los últimos años??)
* ¿Diferentes niveles de compentencia?. Parece que la competencia es similar en los tres sitios: no diferencias sig. para valores de Stand density ySize ratio proportional to distance (ver table 1). 
* ¿Diferencias entre suelos?? En principio no. [@CoboDiaz2017] en un trabajo sobre microbiota del suelo, explora diferencias en un gradiente altitudinal en Cáñar. Analiza 3 sitios: por encima del treline (XZF, el piornal), low altitudinal oak forest (LAF, en la zona baja del robledal) and high altitudinal oak forest (HAF). En su trabajo, LAF está mucho mas bajo, pero HAF está cerca de las parcelas caHigh. Restulados:  
    * tipo de suelo: LAF es Sandy-loam, HAF es loam
    * HAF tuvo los valores mayores de disponibilidad hídrica (% available water tabla s1 en [@CoboDiaz2017]) --> Esto es importante, creo
    * Los tres sitios eran pobres en materia orgánica, pero el HAF dobló los valores de los otros dos sitios
    * No diferencias en contenido de N, C/N ratio similares en HAF, y LAF

* ¿Menor estrés hídrico en zonas mas altas? Puede ser que tengamos menor estrés hídrico en esta zona?? Este robledal, está en la cara sur de SN, y en una ladera con bastante insolación. Algunos trabajos antiguos (de fitosociólogos) hablan de que su presencia aquí se debe a que reciben un aporte extra de humedad procedente de las brisas del mediterráneo, para suplir el mínimo de humedad que necesitan en verano. Quizá también tendríamos que incluir el papel de las acequias. La zona caHigh tiene una acequia muy cerca (y por encima) de donde muestreamos (recordad el roble mas grande, y el mas alto, esta justo en una acequia). No se si es interesante que lo comentemos. 


En un estudio a escala continental, [@Bhuyan2017] encontraron que para la misma especie (estudiaron 850 sites: Fagus sylvatica, Abies alba, Picea abies, Larix decidua, Pinus cembra, P sylvestris, P nigra, Quercus petraea y Q robur) los stands situados a mayores elevaciones fueron menos sensibles a la sequía que los situados a elevaciones inferiores (stands were less drought sensitive at higher elevations compared to lower elevations). 
The effect of elevation was seen clearly in the case of several species where high elevation sites showed greater drought resistance compared to stands at lower elevation in the same climate zone. 



In addition, despite southern sites are very close to each other, we obtained higher values of BAI for the high-elevation site than for the low-elevation site. 

")
```
```{r echo=FALSE, results='asis', eval=show_text}
cat("
### OTRA COSA 

In fact, several authors have raise the need to consider land use y drought de forma conjunta en las environmental y forest management .... (ver Peñuelas et al 2017 y tb Doblas Miranda )

Environmental and forest management policies should take into account all these characteristics of Mediterranean forests and the social, environmental and climatic conditions that are projected for the coming years and decades.
")
```
```{r echo=FALSE, results='asis', eval=show_text}
cat("
#### Climate and tree relationship 

Frase de resultados que tenemos que poner en discussion: 
Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population. Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 7b), specially for the northern population (r > 0.6), ... **which can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014]**.  
")
```
```{r echo=FALSE, results='asis', eval=show_text}
cat("
### ¿Que factor es mas limitante para el crecimiento en el rear-edge de Q. pyrenaica? 
Aquí pueder ser interesante comentar algo de el peso de las variables climáticas en el crecimiento para poblaciones situadas en el borde de distribución (pesa mas la temperatura o la disponibilidad de agua?). 

Para *Q. pyrenaica* moisture availability was reported to be the most limiting factor driving radial growth in Iberian Q. pyrenaica populations [@GeaIzquierdo2014] (Prec hidrológica y SPEI) (ver también Gea-Izquierdo et al. 2015 European Journal of Forest Research). Lo que hemos obtenido aqui (analizando solo el rear edge) también van en esa línea. 

Podríamos complementarlo con lo que le pasa a otras especies en su borde de distribución: por ejemplo en Baza, Herrero et al. 2013, encontraron para Pinus nigra y sylvestris que la temperatura tenía mas peso que la disponibilidad de agua). O también ver algunos de los trabajos de Camarero et al 2013 para el P. nigra en su borde de distribución u otros similares (el de Sanchez-Salguero et al. 2013, 2015) … 

## MAs cosas diferentes 
Otras especies: rear edge 

Evaluar el crecimiento en el rear edge de F. sylvatica 
https://link.springer.com/article/10.1007/s10342-016-0982-7?wt_mc=Internal.Event.1.SEM.ArticleAuthorOnlineFirst 

¿Que limita al crecimiento en el borde sur de su distribucion? --> Ver esto 
http://onlinelibrary.wiley.com/doi/10.1111/j.1365-2486.2006.01250.x/abstract 

Pinus sylvestris (Baza) (Matias et al 2017)
Radial growth was maximal at medium altitude and treeline of the southernmost populations. Temperature was the main factor controlling growth variability along the gradients, although the timing and strength of climatic variables affecting growth shifted with latitude and altitude. 

http://onlinelibrary.wiley.com/doi/10.1111/gcb.13627/full

Idea --> algunos autores comentan que puede existir una alteración en el balance competitivo entre especies en mixed stands. Por ejemplo en Montseny, borde equatorial para F. sylvatica, se está viendo que el F. sylvatica está siendo reemplazado por Q. ilex. ... En SN, el artículo de B. Benito (Climatic Change) habla del remplazo que existirá de Q. pyrenaica por Q. ilex, sin embargo, estamos viendo que los crecimientos son muy grandes y que Q. pyrenaica tiene mucha resiliencia --> entonces que pasa con las predicciones de dichos modelos ???


The presence of *Q. pyrenaica* forests on the southern slopes of Sierra Nevada is partly explained by the additional contribution of humid air from the Mediterranean sea [@Prieto1975]


")
```

En este sentido, sabemos que estos robledales han estado sometido a muchos ciclos de coppiccing, lo que podrían haber reducido su diversidad genética, y por tanto su resiliencia. Pero varios trabajos [@Valbuena2013; @Valbuena2017] han encontrado que esta especie en su límite de distribución sur, muestra unos altos niveles de resiliencia (en este caso genética). 


```{r echo=FALSE, results='asis', eval=show_text}
cat("

### ORDENAR ??? 



.. ver [@VicenteSerrano2013] impacto de las sequías en diferentes zonas (i.e montañas) 
 

# Otras notas: 
* To guide sustainable forest management, forest researchers are asked to provide concrete answers about forest resilience in response to expected climatic trends, and extreme climatic events (Lindner et al., 2014) https://www.sciencedirect.com/science/article/pii/S030147971400379X?via%3Dihub







* Proyeccciones de crecimeinto de para *Q. pyrenaica* sugieren un declive en el crecimiento en las proximas decadas a lo largo de su rango de distribución en la P. Iberica, en diferentes escenarios climáticos [@GeaIzquierdo2013]: Predictions suggest that QUPY productivity would decline in the next decades all along its distributional range in the Iberian Peninsula for all the climate scenarios studied. Este declive debería ser mas dramático en las baja altitudes de las zonas mas calientes del sur de su distribución 

--> vulnerabilidad local (related to rear-edge)
La tendencia de reducción drástica sugerida por el modelo para el dry-edge de la distribución de la especie podría expresar una aumentada vulnerabilidad de los árboles al incremento del estrés hídrico forzado por un clima mas cálido [@GeaIzquierdo2013]. 

Negative trends over recent years and the greater response to moisture availability found at warmer and drier low elevations in the south suggest vulnerability to warming at the local low elevation dry edge of the species's range (e.g. QUPY9). Otros estudios en la zona Mediterranea reportaron un descenso similar en la productividad con un incremento en la vulterabilidad a la sequía ()



Muchos trabajos han analizado la respuesta de la vegetación a las sequías (eg. @Allen2010; @VicenteSerrano2013; @MartinezVilalta2016; @Norman2016), y algunos de ellos han mostrado el efecto de las sequías en especies Mediterráneas (ver por ejemplo Pasho et al. https://www.sciencedirect.com/science/article/pii/S016819231100253X#fig0020; @Camarero2011 combina uso y sequía; añadir alguno mas de GGI). Algunos Some of these species represent southernmost populations in the Mediterranean ambit, which explains their vulnerability to the warmer conditions (Andreu et al., 2007; Sánchez-Salguero et al., 2016).



Case studies that focus on multiple scales - including local scales - and also valuables (Dale et al. 2018 Frontiers) --> Se necesitan casos de estudio que se enfoquen en escalas multiples (incluyendo las locales) y que combinen varias metodologías (Jump et al. ...)


3) rear edge 


- además (Vulnerabilidad de los bosques a los eventos climáticos extremos ...):  

* Existen evidencias que sugieren que muchos bosques son vulnerables a eventos climáticos extremos ... (Zhang) y esto puede ser especialmente relevante para especies situadas en el rear edge (completar) ... 




### Estudios que utilizan RS y TR

* Wu et al. 2017. 
	* En un estudio sobre el efecto de la sequía sobre el crecimiento en bosques, grass y shrub del hemisferio norte, encontraron que los bosques (Deep-rooted forest) exhibit a drought legacy response with reduced growth during up to 4 years despues de un evento extremo de sequía. 

	* Además vieron que los bosques mostraron una stronger drought resilience in forest (i.e. smaller growth reduction after severe drought) (Gazol el al. 2017; Wu et al. 2017). 

	* Esto se puede deber a que acceden a reservas de agua mas profundas (los pastos y los matorrales no pueden acceder a ese agua)




### 
Nosotros hemos obtenido una relación entre la severidad de la sequía y la resiliencia mostrada por los robledales en su rear edge. (ver gráfica )

@Gazol2018 encontró La resiliencia está relacionada con la severidad de la sequía y con la composición del bosque


Rt-NDVI and Rt-TR were positively related with drought intensity 




## 
@Gazol2018 Resistencia a la sequía mayor en la zona norte de españa que en las zonas secas del sur 


Qpyr 
* Valores de Rt-TR aumentaron a lo largo del tiempo, sin embargo Rt-NDVI dismiuyó con el tiempo
* Valores Rc-NDVI tendencia positiva 



")
```


Methodological approach 

comentario: tengo que rematar este apartado.

Vegetation reflects the environmental conditions, and the effects of drought on vegetation can be observed using information from several methodological approaches [see @Norman2016 for a review]. Several works revealed the utility of remote-sensing [*e.g*. @Zhang2013; @AghaKouchak2015] and of the dendrochronology [*e.g.* @Eilmann2012; @Bhuyan2017] to quantify the drought impacts on vegetation at different spatial and temporal scales, but to our knowledge the combination of both approaches is scarce [@VicenteSerrano2013; @VicenteSerrano2016; @Bhuyan2017b; @Wu2018]. Estos trabajos han observado que existe de forma general a positive relationship between vegetation indices derived from remote sensing and annual tree growth [@VicenteSerrano2016; @Bhuyan2017b; @Gazol2018].

Aunque existen algunos trabajos que han utilizado RWI y remote sensing nuestra aproximación (creo) es interesante, sobre todo porque la aplicamos al estudio de poblaciones que están en su límite de distribución (rear-edge) [@Jump2010]. Los estudios sobre poblaciones localizadas en el rear-edge de su distribución requieren de aproximaciones multidisciplinares que analicen la respuesta de las poblaciones al cambio climático y su resiliencia utilizando diferentes aproximaciones metodológicas, por ejemplo combinando dendro y remote sensing [@Jump2010]. 

- Nosotros aqui utilizamos la combinación de técnicas de remote sensing con dendro para analizar la vulnerabilidad de poblaciones de Q. pyrenaica situadas en su reaar edge frente a los eventos de sequía. 

- Hemos encontrado diferencias en cuanto a la resiliencia utilizando RS y RWI. Diferente sensibilidad de satélite frente a los datos de campo. Esto también se ha observado en @Gazol2018, que analizaron la resiliencia a la sequia de 1986, 1994-1995, 1999 y 2005, usando RS y TR. Encontraron que los datos de TR son mas sensibles para la resiliencia del bosque a la sequía que los datos de RS

- De hecho, @Jump2010 recomienda el uso de una aprox combinda de dendro, remote sensing y ground-based assessment para analizar los efectos del cambio global en las poblaciones situadas en el rear edge.


- La utilización de datos de tree-ring nos permite contextualizar los resultados obtenidos con remote sensing, ya que los datos de tree-ring reflejan anomalías en el crecimiento de los árboles (inducidas por clima o por disturbances) durante décadas o siglos [@Babst2017] (... Combined analyses may also allow climate-induced variability in forest growth to be disentangled from that driven by community-level ecological processes. )


- Nota from @Jump2010: Whilst lower resolution data, whether from remote sensing or ground-based monitoring, can inform on widespread regional changes in forest condition, these data are often not adequate for monitoring changes occurring at equatorial range margins owing to their complex distribution and/or topographical variability. In such regions, targeted collection of high-resolution data is necessary in order to identify currently occurring changes and predict the magnitude and spatial distribution of future decline. [@Jump2010]
 
 
 

```{r echo=FALSE, results='asis', eval=show_text}
cat("
# Notas Jump 2010 
Limitación de NDVI. La escala espacial del NDVI limita su uso para estudiar cambios en poblaciones situadas en su rear edge (ver 34 en Jump 2009)

NDVI (o EVI) time series are sufficiently robust to allow comparison of forest growth at seasonal and/or interannual scales. 

Ojo se ha visto que el NDVI responde de forma diferente a la sequía, no solo entre diferentes tipos de bosques, sino también entre poblaciones que difieren en historia de sequía y en riqueza de especies dentro de bosques (ver  26 dentro de Jump 2009) --> Por eso nosotros lo combinamos con el uso de dendro!!

Otra idea --> Limitación de NDVI. La escala espacial del NDVI limita su uso para estudiar cambios en poblaciones situadas en su rear edge (ver 34 en Jump 2009)

NDVI (o EVI) time series are sufficiently robust to allow comparison of forest growth at seasonal and/or interannual scales. 

Ojo se ha visto que el NDVI responde de forma diferente a la sequía, no solo entre diferentes tipos de bosques, sino también entre poblaciones que difieren en historia de sequía y en riqueza de especies dentro de bosques (ver  26 dentro de Jump 2009) --> Por eso nosotros lo combinamos con el uso de dendro!!

El NDVI se ha utilizado para monitorizar la respuesta de bosques frente a sequía (ej.: 24,26,32, 44 en Jump2009 )

Necesidad de Ground-based assessment (p 1974 Jump 2009) -> Aunque utilizemos datos de satélite, siempre es necesario utilizar aproximaciones combinda

Para rear edge poblations es recomendable utilizar aproximaciones combinadas (Jump 2009) donde además de los datos de satélite, se utilicen datos groun based assessment (como por ejemplo la dendro), ya que éstos últimos, además de ser componentes esenciales del forest monitoring, son necesarios para estimar ... (ver esto y completar)

El problema de los datos ground-based assessment, existe un trade-off entre el área total cubierta por una red de monitoreo y su resolución local. --Por tanto también existen limitaciones de las ground-based para estudiar poblaciones en su rear edge. 
")
```



## References 



