---
title: "ms_2_results"
output:  
  md_document:
    toc: true
    variant: markdown_github
bibliography: ../refs/references.bib
csl: ../refs/ecology.csl
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(here)
library(pander)
library(kableExtra)
library(knitr)
```

```{r, echo=FALSE}
show_text <- FALSE
```

# Results

## Time trends in vegetation greenness

The analysis of greenness time trends showed that 78.9 % of the pixels of *Quercus pyrenaica* forests experienced an EVI positive trend for the 2000-2016 period, of which 31.67 % were significant. The lowest values of EVI standardized anomalies were recorded in 2005, being singnificantly lower (-2.285 ± 0.029) than those in 2012 (-0.418 ± 0.029) (LSMEANS, t.ratio = -45.358; p_value < 0.0001), particularly for northern populations (Figure 2). During 2005 drought, *Q. pyrenaica* forests suffered a "browning episode" ($EVI~sa < -1$) (99.36 % and 79.37 % of the pixels for northern and southern populations respectively), yet no changes in greenness were observed in response to the 2012 drought (Appendix S5).

```{r echo=FALSE, results='asis', eval=show_text}
cat("
The analysis of greenness time trends showed that 78.9 % of the pixels of *Quercus pyrenaica* forests experienced an EVI positive trend for the 2000-2016 period, of which 31.67 % were significant trends. The strongest trends were observed in southwestern populations (Figure S2), in which the Cañar site are included Vegetation greenness of *Q. pyrenaica* forests were lower during both the 2005 and 2012 years than the greenness observed for the reference period (Figure 2a). The lowest values for EVI standardized anomalies were recorded in 2005 being singnificantly lower (-2.285 ± 0.029) than those in 2012 (-0.418 ± 0.029) (LSMEANS, t.ratio = -45.358; p_value < 0.0001), particularly for northern populations (Figure 2b). Reduction in annual EVI mean was considerably higher in northern populations than in southern ones during the 2005 drought (Figure 2b). 

According to the standardized anomalies, *Q. pyrenaica* forests suffered a browning episode during 2005 drought event (99.36 % and 79.37 % of the pixels for northern and southern populations respectively), yet no changes in greenness were observed in response to the 2012 drought (Figure S3). 
")
```

```{r echo=FALSE, results='asis', eval=show_text}
cat("
> Cuando exploramos las anomalías (brutas, estandarizadas y normalizadas) observamos valores muy negativos para el año 2005. Sin embargo vemos valores menos negativos para 2005. Tukey posthoc testing (lsmeans package CITAR) was conducted for pairwise comparisons among the slopes and the disturbance years 

> Las anomalías (sa) fueron significativamente menores en 2005 (-2.285 masmenos 0.029) que para 2012 (-0.418 masmenos 0.029), (LSMEANS, t.ratio = -45.358; p_value < 0.0001)
")
```

## Greenness resilience to drought events

*Q. pyrenaica* forest showed significantly lower resistance (*Rt*) to the 2005 drought event than to that in 2012 (Table 3; Figure 3). Southern populations showed significantly higher values of resistance to drought than northern ones (Table 3), except for 2012 where non-significant differences were recorded (Table S1; Figure 3) . 

```{r echo=FALSE, results='asis', eval=show_text}
cat("
> Resistencia
> 
> * Los robledales mostraron menor resistencia a la sequía de 2005 que a la de 2012 [2005: 0.858 (0.853-0.863); 2012: 0.943 (0.939 - 0.947); p <0.0001]
> * Menor resistencia de las poblaciones del Norte a los eventos de sequía que las del Sur [N: 0.883 (0.877-0.889); S: 0.921 (0.918 - 0.925); p <0.0001]
> * La resistencia varió en función de la sequía y de la población. Las poblaciones mostraron una resistencia similar al evento de sequía de 2012 (padj = 0.172), sin embargo las poblaciones del N fueron mucho menos resistentes que las del Sur durante la sequía de 2005  [N: 0.819 (0.814-0.824); S: 0.902 (0.896 - 0.907); p <0.0001]
> * notas:  La sequía de 2005 redujo el EVI medio hasta el 85.8%, mientras que la de 2012 la redujo hasta el 94.3%. La sequía de 2005 redujo el evi medio hasta el 81.9 % en las poblaciones del N, y hasta el 90.2 % en las del S. ... drought reduced growth to xx% of the preceding reference period (ver Pretzsch et al 2013 Plant Biology)
")
```
Recovery (*Rc*) of greenness was significnatly different bewteen drought events and sites (Tables 3). In the 3-year period after the 2005 drought, greenness achieved was 112 % (Rc = 1.12) and after 2012 was 105.7 % (Rc = 1.057) (Table S1). For southern populations, a similar recovery after the 2005 and 2012 drought event was observed (p = 0.2453; Figure 3; Table S1). 

```{r echo=FALSE, results='asis', eval=show_text}
cat("
> Recovery 
> 
> * La recuperación de los robledales fue mayor tras la sequía de 2005 que tras la de 2012 [2005: 1.120 (1.113-1.126); 2012: 1.057 (1.054 - 1.060); p <0.0001]
> * Los robledales de la cara sur mostraron una menor recuperación que los de la cara norte [N: 1.102 (1.096-1.108); S: 1.069 (1.065 - 1.073); p <0.0001]
> * Las poblaciones del sur mostraron una recuperación similar ante la sequía de 2005 y 2012 (p = 0.186), cosa que no ocurrió para las poblaciones N (p < 0.0001), que mostró una recuperación mayor para la sequía de 2005 que para la de 2012 [2005: 1.169 (1.161-1.177); 2012: 1.042 (1.036 - 1.047); p <0.0001]. En 2005, las poblaciones del S mostraron menor recuperación; mientras que en 2012 ocurrión un patrón inverso, mostrando un patron mayor que las del norte. 
")
```

Resilience (*Rs*) was significantly higher for the 2012 drought event than for 2005 (Tables S1, 3), although both values were close to 1 indicating that greenness level was rather similar after each disturbance event (Table S1). The southern populations showed higher resilience values than the northern ones, although these differences were not significant for the 2005 drought event (p = 0.036; Figure 3). 

```{r echo=FALSE, results='asis', eval=show_text}
cat("
> Resilience 
> * La resiliencia de los robledales fue mayor para la sequía de 2012 que para la de 2005 [2005: 0.958 (0.955-0.962); 2012: 0.995 (0.991 - 0.998); p <0.0001]  
> * Los robledales del sur mostraron mayor resiliencia que los del norte [N: 0.970 (0.966-0.974); S: 0.983 (0.980 - 0.986); p <0.0001], aunque para 2005 ambas poblaciones no mostraron diferencias en la resiliencia (padj = 0.152). En 2012 se observó mayor resiliencia en las del S que en la del N (p<0.0001)
")
```

```{r echo=FALSE, results='asis', eval=show_text}
cat("
> Relative Resilience (De esto no he dicho nada)

  * Los robledales mostraron mayor resiliencia relativa a la sequía de 2005 que a la de 2012 [2005: 0.099 (0.095-0.105); 2012: 0.053 (0.050 - 0.056); p <0.0001]  
  * Las poblaciones del sur mostraron menor resiliencia relativa que las del norte [N: 0.086 (0.082-0.092); S: 0.063 (0.060 - 0.066); p <0.0001], debido sobre todo a la diferencia en la resiliencia relativa para las poblaciones del norte entre los dos eventos de sequía (mucha mayor resiliencia relativa en 2012 que en 2005 para las poblaciones del N)
  * Las poblaciones del sur no mostraron diferencias en cuanto a la resiliencia relativa entre los dos eventos de sequía (padj = 0.152)
")
```

## Radial growth trends and climate 

Likely as a consequence of their different management origin (*i.e.* land-use legacy), the trees of the southern population were older than northern ones, particularly for the high elevation site (CA-High) which had bigger and taller trees than the other sites (Tables 1, 2). The competition index was similar among sites but plot basal area was greatest in CA-High (Table 1). Yet the basal area increment (BAI) growth in the highest plot in altitude (CA-High) was the greatest (Figure 4). Trees from this location showed in addition a positive growth trend since the late 1970s. This positive trend was not found in any of the other two locations and the two altitudes sampled in the southern populations showed differences in growth, being more similar CA-Low with growth in the northern population (SJ) than with CA-High (Figure 4; Appendix S11). 

1995 was the lowest pointer year (*i.e.* the lowest growth) since 1950 in all chronologies (Figure 5). For 2005 and 2012 we found a greater reduction of RWI for northern site (SJ) but weaker for southern sites (CA-High and CA-Low).   

```{r echo=FALSE, results='asis', eval=show_text}
cat("
> Relative Resilience (De esto no he dicho nada)

Sorprende que en los sitios del sur (caH y caL), en los últimos años ha habido crecimientos muy grandes. Por ejemplo, en 2010 se observó uno de los mayores valores de RWI de toda la cronología tanto en caH como en caL. Algo similar ocurrió en sj para los años 2003 y 2013. 
")
```

## Resilience of radial tree-growth to drought events

Although no significant differences were observed in the resistance (*Rt*) of oak radial growth between the two drought events (Table 3), the 2005 drought reduced growth more than that of 2012 (Rt = 0.721 and 0.819 respectively) (Table S2). Similar to results for greenness, the northern site, which is under a drier climate, showed resistance values lower than those of the southern site, especially for the 2005 drought event where the growth was reduced to 44.5 % respect to that of the preceding period (Figure 3). 

```{r echo=FALSE, results='asis', eval=show_text}
cat("
> Resistance: 
  
* No diferencias significativas entre la resistencia mostrada por los robledales a los eventos de sequía de 2005 y 2012 (F = 6.0189; p = 0.019). Valores de resistencia menor al evento de 2005. [2005: 0.721 (0.6437-0.7984); 2012: 0.8193 (0.7758 - 0.8628); p = 0.025; trimmed-p = 0.019]

* Diferentes resistencias a los eventos de sequía en función del sitio (p <0.0001). Las localidades del sur (CaLow y CAHigh no mostraron diferencias en cuanto a la resistencia p.adjust = 0.012) mostraron una resistencia mayor a los eventos de sequía que la observada a la localidad del norte [caH: 0.8157 (0.7549 - 0.8764) (a); caL: 0.9209 (0.8834 - 0.9584) (a); SJ: 0.6116 (0.5387 - 0.6846) (b)].

* La interacción también fue significativa. De hecho, si miramos las gráfica, observamos como la resistencia a la sequía de 2005 fue significativamente menor en SJ que la resistancia mostrada, tanto por las otras poblaciones para 2005, como la mostrada por SJ para el año 2012 --¿¿¿ Podemos decir que la sequía de 2005 afectó mucho mas a la población de SJ??? 
")
```

Higher values of recovery (*Rc*) were observed after 2012 than after 2005 drougth events, staying at levels of *Rc* < 1 (Figure 3, Table S2). SJ site (northern aspect) showed significantly higher values of recovery than southern sites (CA-High and CA-Low) (Tables 3 and S2), which even not reach recovery values of Rc = 1 (Figure 3).


```{r echo=FALSE, results='asis', eval=show_text}
cat("
> Recovery 
> * Mayor recuperación del crecimiento de los robledales tras la sequía de 2005 que tras la de 2012 [2005: 0.9462 (0.8794-1.013); 2012: 1.161 (1.081 - 1.24); p < 0.001]
> * Las poblacionees del sur (caH y caL; no diferencias entre ellas) mostraron una menor recuperación que las poblaciones del norte, es decir, las del norte se recuperaron mas rápidamente que las del sur que no llegaron incluso a valores de Rc = 1. 
") 
```

Similar pattern of resilience (*Rs*) values was found for growth than for greenness respect to drought event: significantly higher values of *Rs* for the 2012 drought event than for the 2005 (Table S2; Figure 3). However, no differences were observed between sites (Table 3). For 2005 drought event, *Rs* value of SJ (northern site) was lower than that of southern ones (CA-High and CA-Low), but opposite pattern was found for the 2012 drought event. All values of *Rs* for growth were below 1, except for the SJ site in 2012 (Rs = 1.031). 

When we explored resilience values of growth (*Rs*) for other severe droughts, we found a relation between the resilience and drought severity ($r^2$ = 0.336) (Appendix S6). 


## Climate and tree-growth
Tree-growth exhibited a significant greater response to water availability than to temperatures, as it can be expected for a Mediterranean site and a species located closer to its xeric/dry distribution limit (rear-edge). Cumulative precipitation of the hydrological year was the climatic variable with the highest relationship with growth and influenced similarly and positively the tree-growth for both northern and southern populations (Figure 6a). Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population. Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 6b), specially for the northern population (r > 0.6). Relationships with temperatures were lower than those with moisture variables. Yet there were differences between northern and southern populations: spring maximum temperature was the most significant limitation for the tree-growth of the southern population at both elevations while maximum temperatures of the previous August were the only negative significant for the northern one (Figures 6c and 6d). Remarkably minimum and maximum temperatures of current September positively influenced the tree-growth only in the northern population. No positive relationship with temperature was found for southern trees. 

## Forest disturbances
The analysis of growth changes revealed differences forest history between sites (Figures 7 and 8). Northern site (SJ) showed two release events ($GC$ > 50 %) detected at stand-wise scale (occurring in more than 50 % of sampled trees): the first during the 1940 decade and the second in the period 1994-2001. These periods alternate with periods of supression. Southern sites (CA-High and CA-Low) showed some weak suppression episodes, but not in the last 50 years. 



```{r echo=FALSE, results='asis', eval=show_text}
cat("

En las poblaciones del norte osbervamos dos eventos de PGC por encima del 25 e incluso el 50 %. El primero de ellos desde 1940 hasta 1950, mientras que el segundo entre 1994 y 2001. Estos periodos se alternan con periodos de NGC: previo 1940, 1960 - 1991, y quizá aunque no lo vemos en al actualidad (?? duda. Esto es especular un poco). Esto responde creo que a: 

(i) actividad minera y tratamientos forestales. La actividad minera en la zona fue intensa (aunque a veces intermitente, entre 1890 y 1957 (cuando se clausuró); y previamente desde 1858 tras los trabajos de Amalio Maestre). Hay costancia de la existencia de hornos de fundición de mineral en alguna de las minas (La Probadora).
(ii) actividades forestales:
* Se conoce la existencia de un proyecto redactado en 1943 de repoblación forestal de la cabecera del río Genil (he pedido los mapas).
* Ver actividades forestales (bd proyectos de gestión)

En el trabajo de [@GeaIzquierdo2014] se observa una liberación posterior a 1950 (QPUY9) (entre 1950 y 1960 aprox.) y luego una supresión hasta mediados de 1990. Esto coincide con lo que hemos observado en nuestra chrono. 

Respecto a las poblaciones del sur, destaca en la caH una primera liberación en torno a 1829-1830 y 1837-1840; y luego predominan los eventos de supresión que se alternan con algunos de liberación (muy débiles) hastas 1950. Posteriormente se alternan liberaciones-supresiones débiles. Destaca una liberación en torno a 2000-2003 (relación con clima??? posterior a sequía 1999???)

En caL, observamos un periodod de limitación hasta 1885 aprox. Luego se alternan al igual que antes periodos de liberación con otos de supresión con señales más fuertes que en caL, y al igual que antes destaca con mas intensidad si cabe, una liberación entre 2000 y 2003 (antrópico o climática ??)

Duda: ocurre en los dos sitios caL y caH, entonces, es aumento de crecimiento debido a actuación forestal? o debido a clima?? 

[@GeaIzquierdo2014] encontró un pico de disturbance en torno a 1860 en QUPY10 (cañar) sugiriendo que la baja densidad en esos sitios fue presumible el resultado de un thinning (aclareo). El patrón de supresión que observa para este sitio va en la linea de lo que nosotros encontramos (supresiones que alcanzan a veces el 25 % y otras no)


En cuanto a la señal a nivel de sitio, para el Norte, vemos que tanto las supresiones como las liberaciones se observan para mas del 50 % de los árboles. En el caso del sur, vemos que algunas supresiones se observan para mas del 50 % de los árboles, mientras que las liberaciones observadas en torno a los 2000 se observan para mas del 50 % de los árboles 
")
```


