---
output: 
  word_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
```

```{r load_pkg}
library("tidyverse")
library("here")
library("ggrepel")
library("scales")
library("grid")
library("gridExtra")
library("patchwork")
library("ggcorrplot")
library("stringr")
library("pander")
library("lemon")
```


```{r data}
ltdf <- read.table(file=here::here("/data_raw/meteo", "long_term_data_rediam_data.txt"), header=T, sep=';', dec=',')

source(here::here("/script/R/", "hydro_year.R"))

ltdf <- ltdf %>% 
  dplyr::select(cod = INDICATIVO, VARIABLE, FECHA, VALOR) %>% 
  dplyr::mutate(fecha = as.Date(FECHA, format='%d/%m/%Y'),
                month = lubridate::month(fecha),
                hmonth = ifelse(month <= 8, month + 4, month -8), 
                hyear = hydro_year(fecha),
                hyear_f = paste0(hyear-1,'-', hyear))


# We used only the 5514 station (the most long term)
# Filter by station and prec 
# Filter from 1950 
st5514 <- ltdf %>% 
  filter(VARIABLE == 'PM1') %>% 
  filter(cod == '5514') %>% 
  filter(hyear >= 1950) %>% 
  dplyr::select(fecha, month, hmonth, hyear, hyear_f, pre=VALOR)

st5514_pro <- st5514 %>% 
  group_by(hyear_f, hmonth) %>% 
  dplyr::summarise(pre = sum(pre)) %>%
  dplyr::mutate(csum = cumsum(pre)) %>% as.data.frame()

df <- st5514_pro  %>% 
  filter(!(hyear_f %in% c('2004-2005', '2011-2012')))

```

```{r plot_cummulatedPrec}

size_line <- .25 

cp <- ggplot(df, aes(x=as.factor(hmonth), y=csum)) + 
  geom_boxplot(aes(x=as.factor(hmonth)), size = 0.25, outlier.size = 0.5) +  
  geom_smooth(aes(group=1), se=TRUE, level=0.95, colour='black', size = size_line) + 
  geom_line(data=subset(st5514_pro, hyear_f == '2004-2005'),
            aes(x=as.factor(hmonth), y=csum, group=1), colour='blue', size = size_line) +
  geom_line(data=subset(st5514_pro, hyear_f == '2011-2012'), 
            aes(x=as.factor(hmonth), y=csum, group=1),
            colour='red', size = size_line) +
  theme_bw() + 
  theme(strip.background = element_rect(fill = NA, color = "black"),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5, size=7),
        axis.text.y = element_text(size=7),
        axis.title.y = element_text(size = 7)) + 
  ylab("Cumulative \n Precipitation (mm)") + xlab('') + 
  # ggtitle('Granada Base Aerea, 1950-2015') + 
  scale_x_discrete(labels=c("1"="Sep", "2"="Oct", "3"="Nov", "4"="Dec", "5"="Jan", "6"="Feb",
                            "7"="Mar", "8"="Apr", "9"="May", "10"="Jun", "11"="Jul", "12"="Aug")) +
  annotate("text", label = "2004-2005", x = 2, y = 420, size = 2, colour = "blue") + 
  annotate("text", label = "2011-2012", x = 2, y = 520, size = 2, colour = "red") + 
  annotate("text", label = "1950-2015", x = 2, y = 620, size = 2, colour = "black") +
  theme(panel.grid = element_blank(), 
        axis.title.y = element_text(hjust = 0.5),
        plot.background = element_blank())
```


```{r data_prec}
prec_files <- list.files(here::here('data_raw/meteo/prec_subclima/'), pattern="*.csv")

dfpre <- c() 

for (i in prec_files){
  
  auxdf <- read.csv(here::here('data_raw/meteo/prec_subclima', i), header = TRUE, sep=";")
  auxdf$code <- gsub(x=i, pattern = ".csv", replacement = "")
  
  dfpre <- rbind(dfpre, auxdf)
} 


dfpre$date <- as.Date(dfpre$date, format = "%d/%m/%Y")

dfpre <- dfpre %>% 
  dplyr::mutate(fecha = as.Date(date, format='%d/%m/%Y'),
                month = lubridate::month(fecha),
                hmonth = ifelse(month <= 8, month + 4, month -8), 
                hyear = hydro_year(fecha),
                hyear_f = paste0(hyear-1,'-', hyear))


aux <- dfpre %>% 
  group_by(hyear_f, code) %>% 
  dplyr::mutate(csum = cumsum(prec)) %>% 
  filter(hmonth == 12) %>% 
  dplyr::select(-month, -prec, -date, -fecha) %>% as.data.frame()


presn <- aux %>% group_by(hyear) %>% 
  dplyr::summarise(
    csum_mean = mean(csum, na.rm = TRUE),
    csum_sd = sd(csum, na.rm = TRUE), 
    csum_se = csum_sd/sqrt(length(csum_sd)), 
    n = n())  

meandf <- mean(presn$csum_mean)
sddf <- sd(presn$csum_mean)
```

```{r plot_prec}
yeah <- presn %>% filter(csum_mean < (meandf - 1*sddf)) %>% 
  mutate(pos = csum_mean - 50) %>%  as.data.frame()
  
prec_hydrol_sn <- ggplot(presn, aes(x=hyear, y=csum_mean)) + 
  geom_hline(yintercept = meandf) + 
  geom_hline(yintercept = meandf + 2*sddf, color ='blue', linetype = 'dashed') +
  geom_hline(yintercept = meandf - 2*sddf, color ='red', linetype = 'dashed') +
  geom_hline(yintercept = meandf + 1*sddf, color ='blue', linetype = 'dotted') +
  geom_hline(yintercept = meandf  - 1*sddf, color ='red', linetype = 'dotted') +
  geom_errorbar(aes(ymin = csum_mean - csum_se, ymax= csum_mean + csum_se), width=.1, colour='gray') +
  geom_point(size=2) + theme_bw() + 
  geom_line(size=.25, colour = 'black') + 
  scale_x_continuous(breaks=seq(1950,2015, by=5)) +
  scale_y_continuous(breaks=seq(0,1500, by=250)) + 
  geom_text_repel(
    data = yeah,
    aes(label = hyear), 
    size = 2.5, 
    alpha = .5,
    colour = "black",
    segment.size = 0.1, 
    segment.alpha = 0.5,
    point.padding=unit(.7,'lines'),
    nudge_y = -yeah$csum_mean,
    direction = 'x'
    ) +
  theme(panel.grid = element_blank(),
        axis.title = element_text(size = 9),
        axis.text.y = element_text(size = 8)) +
  ylab('Cumulative Precipitation (mm)') + xlab('Year') +
  expand_limits(y=c(0,2000)) 
```

```{r plot_all}
p <- prec_hydrol_sn + annotation_custom(ggplotGrob(cp), xmin = 1990, xmax = Inf, ymin=1050)
```

**Figure S1.** Temporal evolution of cumulative precipitation (hydrological year) during the period 1950-2017. Points represent mean and errorbars standard error. *Black* line indicates mean for all period. *Red* lines represent -1 and -2 standard deviation (*dotted* and *dashed* lines respectively). *Blue* lines represent +1 and +2 standard deviation (*dotted* and *dashed* lines respectively). Years with average values below -1SD are labelled. Data from 28 meteorological stations distributed around Sierra Nevada area (from National Spanish Meteorological Services, AEMET). ***Inset plot***: cumulative precipitation during the hydrological years 2004-2005 (*blue line*) and 2011-2012 (*red line*). The boxplot representing the average from 1950-2015 period. Data from meteorological station Granada, Base Aérea. 

```{r plot_out, fig.width=7, dpi = 300, out.width = 0.8} 
p
```

