---
output:
  word_document
bibliography: ../../refs/references.bib
csl: ../../refs/ecology.csl
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
```


```{r load_pkg}
library("tidyverse")
library("here")
library("lubridate")
library("anytime")
library("rdryad")
library("scales")
```

**Figure S7.** EVI annual profile (average of the period 2000-2016) for *Q. pyrenaica* forests in Sierra Nevada and drought events. Horizontal bars correspond to the most severe droughts for Sierra Nevada since 1900 (computed as in Table S3). Their position indicate the start and end months of each drought event. Bars length show the duration of the drought event (number of consecutive months with SPEI lower than -1.28, see [@Pascoa2017]). 

```{r preparedata}
# Datos de iv 
iv <- read.csv(here::here("data/evi/iv_composite.csv"), header = TRUE, sep = ',')

evi_profile_dat <- iv %>% 
  group_by(composite) %>% 
  dplyr::summarise(mean=mean(evi, na.rm=T),
            sd = sd(evi, na.rm=T),
            se = sd/sqrt(length(evi))) %>% 
  mutate(composite_dates = 
           plyr::mapvalues(composite,
                           c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23),
                           c('01-01','01-17','02-02','02-18','03-06','03-22','04-07','04-23',
                             '05-09','05-25','06-10','06-26','07-12','07-28','08-13','08-29',
                             '09-14','09-30','10-16','11-01','11-17','12-03','12-19'))) %>%
  mutate(cd = as.Date(composite_dates, format = '%m-%d'))


# datos de Drought
dy <- read.csv(file=here::here('data/sequias', 'severe_spei12.csv'), header=TRUE)

# Convertir en datos de fecha inicio y fecha fin y convertir en falsa fecha

dy <- dy %>% 
  filter(d_duration > 2) %>%  
  separate(rangeDate, c("init", "end"), sep=" - ", remove = FALSE) %>% 
  mutate(miny = ifelse(minyear %in% c(1913,1998), 2018, 2019), 
         maxy = ifelse(maxyear == 1946, 2020, 2019),
         iD = anydate(paste(miny, init, sep="-")),
         eD = anydate(paste(maxy, end, sep="-"))) %>% 
  mutate(r = rank(d_severity, ties.method = "first")) %>% 
  mutate(mean = rescale(r, to =c(0.2, 0.33))) %>% 
  mutate(year = ifelse(minyear == maxyear, minyear, paste(minyear,maxyear, sep="-")))

```

```{r plotEVI, eval = FALSE, echo = FALSE}
micolor <- '#455883'

evi_profile <- ggplot(evi_profile_dat, aes(cd, y=mean)) + 
  geom_errorbar(aes(ymin = mean - 10*se, ymax= mean + 10*se), width=4, colour=micolor, size=.8) + 
  #geom_errorbar(aes(ymin = mean - sd, ymax= mean + sd), width=4, colour='black') +
  geom_line(colour=micolor, size=.8) + 
  geom_point(size=3,colour=micolor) +
  geom_point(size=1.5, colour='white')+
  scale_x_date(labels = function(x) format(x, "%b"),
               breaks = date_breaks('month')) + 
  ylab('EVI') + xlab('Date') + 
  theme_bw() +
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank()) +
  theme_classic()
#format(x, "%d-%b") 

combinado <- evi_profile + 
  geom_segment(data = dy, aes(y = mean, x= iD, xend =eD, yend = mean, colour = d_severity), 
               size = 8, alpha = 0.7) +
  geom_text(data = dy, aes(label = year, x = (iD + floor(eD-iD)/2))) +
  scale_colour_gradient(low = "orange", high = "red") + 
  labs(colour = "Drought Severity")
```

```{r}
micolor <- '#455883'

# PLOT EVI + DROUGHT 
combinado2 <- ggplot(evi_profile_dat, aes(cd, y=mean)) + 
  scale_x_date(labels = function(x) format(x, "%b"), breaks = date_breaks('month')) + 
  theme_bw() +
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank()) +
  theme_classic() +
  geom_segment(data = dy, aes(y = mean, x= iD, xend =eD, yend = mean, colour = d_severity), 
               size = 8, alpha = 0.7) +
  geom_text(data = dy, aes(label = year, x = (iD + floor(eD-iD)/2))) +
  scale_colour_gradient(low = "orange", high = "brown") + 
  labs(colour = "Drought Severity") +
  geom_errorbar(aes(ymin = mean - 10*se, ymax= mean + 10*se), width=4, colour=micolor, size=.8) + 
  geom_line(colour=micolor, size=.8) + 
  geom_point(size=3,colour=micolor) +
  geom_point(size=1.5, colour='white') + 
  ylab('EVI') + xlab('') 


# Get data from Dryad 
# Peres de Lis 2016 et al. https://datadryad.org/resource/doi:10.5061/dryad.1cn19
# x <- dryad_files('10.5061/dryad.1cn19')
```


```{r plot_out, fig.height=5, dpi = 300, out.width = 0.8}
combinado2
```





