---
output: 
  word_document
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE)
```


```{r load_pkg}
library("tidyverse")
library("here")
library("ggrepel")
library("scales")

library("grid")
library("gridExtra")
library("patchwork")
library("ggcorrplot")
library("stringr")
library("pander")
library("lemon")
source(here::here('/script/R/custom_functions.R'))
```


```{r data}
rtmax <- read.csv(here("data/rwi_climate", "rtmax_selected.csv"), header = TRUE, sep = ',') %>% mutate(v_name = 'T max')
rtmin <- read.csv(here("data/rwi_climate", "rtmin_selected.csv"), header = TRUE, sep = ',') %>% mutate(v_name = 'T min')
rsequia <- read.csv(here("data/rwi_climate", "rsequia_selected.csv"), header = TRUE, sep = ',') %>% mutate(v_name ='SPEI')
rprec <- read.csv(here("data/rwi_climate", "rprec_selected.csv"), header = TRUE, sep = ',') %>% mutate(v_name = 'Prec')

df <- rbind(rtmax, rtmin, rsequia, rprec)

label_var <- c("Prec" = "(a) Prec", "SPEI" = "(b) SPEI", 
               "T max" = "(c) T max", "T min" = "(d) T min") 

label_site <- c("SJ" = "SJ",
                "caH" = "CA-High",
                "caL" = "CA-Low")
```

```{r plot_options}
theme_ring_climate <- list(
  geom_hline(yintercept = 0, colour="gray"), 
  geom_bar(aes(fill=site), stat="identity", position = position_dodge2(padding = 0.1)),
  geom_text(aes(label = sig, x = id, y = ifelse(coef > 0, (coef + 0.02), (coef - 0.06))), 
            position = position_dodge(.9), size = 3), 
  theme_bw() +
  theme(axis.title.x = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_line(size=.5),
        axis.text.x.top = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 7),
        axis.text.x.bottom = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
        axis.text.y = element_text(size = 7), 
        axis.title.y = element_text(size = 8),
        legend.text = element_text(size = 7), 
        legend.box.spacing = unit(0.01, "cm")),
  scale_fill_manual(values = colores_den, labels = label_site)
) 
#c("#B2DF8A","#1F78B4","#A6CEE3")
```

```{r plot_tmax}
p_tmax <- df %>% filter(v_name == 'T max') %>% 
  ggplot(aes(x=id, y=coef, group = site)) + 
  theme_ring_climate + 
scale_y_continuous(limits = c(-0.4, 0.8), 
                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4, 0.6, 0.8), 
                     sec.axis = dup_axis(name = NULL)) +
  scale_x_continuous(breaks = unique(df$id), labels = unique(df$month_name), position = 'bottom') + 
  ylab('Correlation') + 
  annotate("text", x = 2, y = 0.35, label = "T max", fontface =2, size = 3.5)
```

```{r plot_tmin}
p_tmin <- df %>% filter(v_name == 'T min') %>% 
  ggplot(aes(x=id, y=coef, group = site)) + 
  theme_ring_climate + 
  scale_y_continuous(limits = c(-0.4, 0.8), 
                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4, 0.6, 0.8), 
                     sec.axis = dup_axis(name = NULL)) +
  scale_x_continuous(breaks = unique(df$id), labels = unique(df$month_name), position = 'top') + 
  ylab('Correlation') +
  annotate("text", x = 2, y = 0.35, label = "T min", fontface =2, size = 3.5)
```


```{r plot_prec_spei}
# prec-spei

dfspei <- df %>% 
  filter(v_name == 'SPEI') %>% 
  filter(month_name %in% c('Hydrol', 'Winter','Spring','Summer','Autumn')) %>% 
  # truquillo para poner juntas las dos graficas 
  mutate(id = id + 6)

dfprec <- df %>% filter(v_name == 'Prec')  

# Joins df
dfprec_spei <- bind_rows(dfspei, dfprec)

# solucion para poner juntas prec + spei 
labelitas <- unique(dfprec_spei[c('id','month_name','v_name')])

p_prec_spei <- dfprec_spei %>% 
  ggplot(aes(x=id, y=coef, group = site)) + 
  geom_vline(xintercept=20) +
  theme_ring_climate + 
  scale_y_continuous(limits = c(-0.4, 0.8), 
                     breaks = c(-0.4, -0.2, 0, 0.2, 0.4, 0.6, 0.8), 
                     sec.axis = dup_axis(name = NULL)) +
  scale_x_continuous(breaks = labelitas$id, labels = labelitas$month_name, position = 'top') + 
  ylab('Correlation') +
  annotate("text", x = 2, y = 0.75, label = "Prec", fontface =2, size = 3.5) + 
  annotate("text", x = 23, y = 0.75, label = "SPEI", fontface =2, size = 3.5)
```

```{r plot_all}
ptemperatura <- p_tmin + theme(legend.position="none") + p_tmax + theme(legend.position="bottom") + plot_layout(ncol=1) 

ring_climate <- p_prec_spei + theme(legend.position="none") + ptemperatura + plot_layout(ncol=1, heights = c(2,3), tag_level = 'keep')  + plot_annotation(tag_levels = c('a','b')) & theme(plot.tag = element_text(size = 12, face = 2)) &
theme(plot.margin = margin(t=0,r=0,b=-2,l=0, "pt"))
```

**Figure 8.** Correlation coefficients obtained by relating tree-ring residual chronologies (RWI) of *Q. pyrenaica* and monthly climatic data: precipitation and 6-month SPEI (a), minimun (b) and maximun (c) temperatures. *green* bars: northern site (SJ); *light blue* bars: low-elevation southern site (CA-Low); and *dark blue* bars: high-elevation shouthern site (CA-High). Asteriks indicate significant ($P < 0.05$) correlation coefficients. 

```{r plot_out, fig.height=5, dpi = 300, out.width = 0.8} 
ring_climate
```


```{r}
# SAVE as eps 
saveImage(ring_climate, path="./man/figures/svg/", dpi=300, w=25, h=14, fig= "fig_8", u = "cm") 
```
