---
title: "Relict deciduous oaks show high resilience to drought along a climatic gradient in the rear-edge despite strong land-use legacies "
output:
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    highlight: null
    reference_docx: ../../templates/template_ms.docx
bibliography: ../../refs/references.bib
csl: ../../refs/ecosystems.csl
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(here)
library(pander)
library(kableExtra)
library(knitr)
```

```{r setup, echo=FALSE}
show_text <- FALSE
```

## **Short Title**
Relict oaks resilience at their rear-edge

## Authors [^\*^] 
A.J. Perez-Luque^1,2,a^; G. Gea-Izquierdo^3,b^ and Zamora, R.^1,2,c^ 

^1^ Instituto Interuniversitario de Investigación del Sistema Tierra en Andalucía (CEAMA), Universidad de Granada. Avda. del Mediterráneo s/n, E-18006 Granada, Spain 

^2^ Grupo de Ecología Terrestre, Departamento de Ecología, Facultad de Ciencias, Universidad de Granada, Avda. Fuentenueva s/n, E-18071 Granada, Spain

^3^ INIA-CIFOR. Ctra. La Coruña km 7.5. E-28040 Madrid, Spain 

^a^ ajperez@ugr.es

^b^ gea.guillermo@inia.es

^c^ rzamora@ugr.es

[^\*^]: Author Contributions. AJPL, GGI and RZ conceived the study, conducted field work and collected the data. AJPL and GGI conducted lab work and led article preparation. AJPL analyzed data and led the writing of the paper. GGI and RZ contributed in the writing process. All authors contributed to the drafts and gave final approval for publication.  


## Manuscript highlights
* *Quercus pyrenaica* rear-edge forests showed high resilience at tree and stand-levels
* The responses to climate and resilience followed a water-stress gradient
* Trees and stand expressed high sensitivity to drought and land-use legacies

## Abstract

* Global change challenges ecosystems in dry locations transformed by intensive human use. We analyzed resilience to drought of relict Mediterranean *Quercus pyrenaica* Willd. populations in the Southern Iberian Peninsula. Forest resilience to drought was analyzed in relation to historical records of land-use using dendroecological growth and greenness (EVI) as proxies for secondary and primary growth. 

* The response to climate and resilience followed a water-stress gradient. Primary and secondary growth of more xeric populations (*i.e.* Northern aspect and low elevations) was less resilient to drought than that at more mesic sites. In addition, a positive growth trend matching the warming climate since the late 1970s was particularly evident in mesic high-elevation stands, but absent in the most xeric site. Trees were highly sensitivity to drought and both primary and secondary growth expressed high resilience to drought events in the short- and the long-term. The growth trends reflected a strong influence of land-use legacies (*e.g.* firewood extraction) in the current forest structure. 

* The species showed high resilience to drought despite being relict and at its rear-edge. Yet, the trees exhibited a variable response to the changing climate along a very short climatic gradient. The trees were more sensitive to water availability at most xeric sites whereas only the coldest and more humid stands exhibited a recent positive growth trend. Stand dynamics at the most xeric locations (*i.e.* the altitudinal rear-edge) needs to be monitored under future expected climatic scenarios bearing in mind that land-use legacies play a key role in stand dynamics.

* Despite being considered relict stands at the dry rear-edge for the species, the trees showed high resilience in response to drought. Yet, there were a divergent response to a changing climate along very short climatic gradients. The oaks were more sensitive to water availability expressed at the most xeric sites whereas they exhibited positive growth trends only in the coldest and more humid stands. These results suggest that stand dynamics, particularly at the lowest elevations (*i.e.* the altitudinal rear-edge), need to be monitored under future expected drier and warmer climatic scenarios bearing in mind that management legacies likely play a key role in current stand dynamics.

## Keywords
extreme drought, resilience, rear-edge, *Quercus pyrenaica*, tree-growth, dendrocronology, remote-sensing



# Introduction

Intense drought is one of the most frequent and widespread climatic extremes affecting ecological systems worldwide and is considered a key feature of Mediterranean climate [@Dai2011]. There has been an increase in the frequency and severity of drought events in the last decades, along with a trend towards drier summers particularly for southern Europe [@VicenteSerrano2014; @Stagge2017]. This is especially critical, since the climate change projections for the Mediterranean area forecast that climate extreme events, like droughts, will become more frequent and severe [@Spinoni2017a].

Drought causes significant changes in ecosystem productivity and water dynamics that can severely affect forests through changes in plant physiology, phenology and carbon allocation. The assessment of the drought impacts on ecosystems has gained attention in the last decades [@Allen2010; @Clark2016], and the functional response of vegetation to drought has been analyzed at global and local scales [*e.g.*, @VicenteSerrano2013; see @MartinezVilalta2016 for a revision].  

Ecosystem responses to drought are influenced by other drivers of environmental changes [@Penuelas2017]. In a global-change context, the impacts of drought on vegetation need to be evaluated considering the interactions with other drivers of change [@DoblasMiranda2017]. This is especially relevant for areas with a long history of landscape modification, such us the Mediterranean region, where land-use change plays a significant role in current ecosystem ecology [@Navarro2013], considering also that the interactions of these two factors, namely drought and land-use change, are crucial for Mediterranean forests [@DoblasMiranda2017].

The impacts of drought are also especially relevant for populations located in the rear edge (either latitudinal or altitudinal along climatic gradients) of their distribution, where climatic conditions often corresponds to the species-limits and the populations are likely to be more sensitive to small variations on climate [@Hampe2005]. The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010]. It becomes even more important for relict populations driven by historical land-uses, which can harbour high levels of intraspecific genetic diversity [@Valbuena2013].

Many studies have analyzed the effects of drought on Mediterranean tree species representing the southernmost limit of their distribution [*e.g.*, @SanchezSalguero2012; @Dorado2017b; @Caminero2018]. Some of these studies focused on the resilience to drought of rear-edge populations [*e.g.*, @Herrero2014]. Resilience is the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance [@Holling1973]. The assessment of resilience to climate disturbances like droughts provides critical information about the capacity of the forests to maintain their structure and to continue providing valuable ecosystem services. This is particularly relevant for populations located in the rear-edge of their current distribution, especially when they are located in Mediterranean mountains.

The resilience to drought of species is better assessed when several aproacches like remote sensing and dendrocronology are combined simultaneously. Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality [@Fritts1976; @Bhuyan2017] and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976]. The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological or elevational gradients [*e.g.* @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018]. Remote sensing can be used for studying droughts impacts on ecosystems [*e.g.* @Zhang2013]. Vegetation indices obtained from satellite are used as a proxy for assessing vegetation functioning and aboveground net primary production [@AlcarazSegura2009], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013]. Drought impacts on vegetation using remote sensing have been analyzed at different scales. For instance, a global-scale study showed a stronger influence of drought for drier biomes [@VicenteSerrano2013] whereas for Mediterranean area, studies at regional and local scales found a large spatial and seasonal variability of the effects of drought on vegetation [@Gouveia2017]. 

Vegetation responses to drought requires of integrative and multidisciplinary strategies for assessing the impacts of disturbances on forest at several scales [@Hartman2018]. Several works assessed vegetation responses in term of resistance and resilience to drought events highlithg a stronger reponse of tree-ring (tree-level) to drought variability compared to the greenness (ecosystem-level) [*e.g.* @PenaGallardo2018; @Gazol2018]. Nonetheless, is crucial to know if the responses at the tree level differ from responses at the ecosystem level, particularly in rear edge populations. 

In this work we used remote sensing information and dendroecological methods to evaluate the drought impacts in both canopy greenness (as a proxy to primary growth) and tree-radial growth of a Mediterranean tree species (*Quercus pyrenaica* Willd.) on southern relict forests at the rear-edge of species distribution and where species performance is considered to be threatened by climate change. We also assessed the resilience of these forests both to several extreme drought episodes and to climate change in the long-term in the last decades. Our main hypothesis is that these relict populations driven by historical land-use at their present climatic (either altitudinal or latitudinal) rear-edge are particularly vulnerable to climate change, and hence they will show lower-values of resilience after several extreme drought only at the dry edge of climatic gradients. To address this hyphotesis our specifics aims were: (*i*) To quantify how two recent extreme drought events influenced primary and secondary growth of *Q. pyrenaica* forests in their rear edge; (*ii*) to analyze the long-term resilience of these forests to extreme drought events using time-series of radial growth; (*iii*) and to explore differences in the resilience metrics between populations located in contrasting ecological conditions within the rear edge of the distribution of this species to detect vulnerability to climate change along climatic gradients. To achieve goal (*iii*) within the region of study we assessed whether the effect of aspect and microclimatic conditions expressed in northern and southern populations of Pyrenean oak forests differ in their resistance, resilience and recovery to extreme drought events.

# Materials and methods 

## Tree species and study site 

Pyrenean oak (*Quercus pyrenaica* Willd.) forests extend through south-western France and the Iberian Peninsula reaching its southern limit in mountains areas of northern Morocco [@Franco1990]. In the Iberian Peninsula these forests live on siliceous soils under meso-supramediterranean and mesotemperate areas and subhumid, humid and hyperhumid ombroclimate. *Q. pyrenaica* requires over 650 mm of annual precipitation and a summer minimal precipitation between 100 and 200 mm. Summer rainfall is a key factor for the species, which is considered to have low drought tolerance compared to genuine evergreen Mediterranean taxa [@delRio2007].

The forests of this species reaches one of their southernmost European limit in Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations between 860 m and 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August). There are eight Pyrenean oak patches (2400 Has) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys. Sierra Nevada mountain is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002]. In this mountain region, *Q.* *pyrenaica* woodlands represent a rear edge of their habitat distribution, which is important for determining habitat responses to expected climate change [@Hampe2005]. They are the richest vascular-plant forest formation of Sierra Nevada, sheltering several endemic and endangered plant species [@Loriteetal2008ABG]. These relictic forests undergone intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2017]. Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" [@Vivero2000] and they are expected to suffer the impact of climate change, reducting its suitable habitats in the rear-edge [@GeaIzquierdo2013]. 

## Drought episodes

During the second half of the twentieth century the Iberian Peninsula suffered several extreme drought episodes (*e.g*. 1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014].The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Pascoa2017] and they were characterized as extreme drought in our climatic data (Figures S1-S2; Table S3). We focused on these two drought events because they were included in the period where there is availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events and analyze forest resilience to drought in a longer term (see Table S3). The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Spinoni2015. We used SPEI-12 months scale for all 0.5º grid cells covering Sierra Nevada obtained from Global SPEI Database (http://spei.csic.es/database.html). A severe drought event starts in the month when SPEI falls below the threshold of -1.28 [@Spinoni2017a; @Pascoa2017]. A drought event is considered only when SPEI values are below the threshold for at least two consecutive months. Several indicators were computed for each drought event: ***duration*** as the number of consecutive months with the SPEI lower than a certain threshold; ***severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event; the ***intensity*** and the ***Lowest SPEI*** refer to the mean and lowest value of SPEI respectively during the drought event.  

## Greenness data 

Vegetation greenness of *Q. pyrenaica* was characterized by means of the *Enhanced Vegetation Index* (EVI), derived from MOD13Q1 product of the MODIS (*Moderate Resolution Imaging Spectroradiometer*) sensor. MODIS EVI Data (Collection 6) were obtanied for the period 2000 - 2016. EVI data consits of 16-day maximum value composite images (23 per year) of the EVI value with a spatial resolution of 250 m x 250 m. We selected the pixels covering the distribution of *Q. pyrenaica* forests in Sierra Nevada (*n* = 928 pixels). Values affected by high content of aerosols, clouds, snow and shadows, were filtered out following recommendations for mountain regions [@Reyes2015]. 
 
As a surrogate of mean annual primary production, annual mean EVI ($EVI_{mean}$) value was computed for each pixel for the period 2000 - 2016. To explore the effect of drought on greenness, the EVI standardized anomaly ($EVI~sa$) was computed pixel-by-pixel, since it minimizes biases in the evaluation of anomalies, providing more information about the magnitude of the anomalies [@Samanta2012]. For each pixel, an annual EVI value was obtained by averaging EVI valid values. Then, the standardized anomaly was computed as: 

$$\mathrm{EVI_{sa,\mathit{i}}}= \frac{\mathrm{EVI_{mean,\mathit{i}}-EVI_{mean,ref}}}{\sigma_{\mathrm{ref}}}$$
where $\mathrm{EVI_{sa,\mathit{i}}}$ is the EVI standardized anomaly for year $i$;  $\mathrm{EVI_{mean,\mathit{i}}}$ the annual mean value of EVI for year $i$; $\mathrm{EVI_{mean,ref}}$ the average of the annual EVI values for the period of reference 2000-2016 (all except $i$ year); and $\sigma_{\mathrm{ref}}$ the standard deviation for the reference period. Each pixel was categorized according the EVI standardized anomalies as "greening" ($EVI~sa > 1$), "browning" ($EVI~sa <- 1$) or "no changes" ($-1 > EVI~sa > 1$)[@Samanta2012]. 

Rather than other vegetation indices like the NDVI, $EVI_{mean}$ was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual ($r$ = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88). For exploratory purposes, temporal trends of $EVI_{mean}$ were explored at a pixel scale, using Mann–Kendall nonparametric test.  

## Field sampling and dendrochronological methods 

Tree sampling was carried out during autumn 2016. Trees were sampled at two locations in contrasting N-S slopes of Sierra Nevada: San Juan (SJ), a xeric site located at northern aspect; and Cáñar (CA), a humid but warmer site located at the southern aspect (Figure 1; Table 1). For the southern site two elevations were sampled: CA-Low and CA-High, which constitutes the current altitudinal rear-edge (CA-Low) and tree-line (CA-High), respectively in the sampled slope. Sampling sites follow a moisture gradient: SJ < CA-Low < Ca-High (Table 1). All the sites were oak monospecific and representatives of the population clusters identified for the species in this mountain range [@PerezLuque2015]. In each site between 15 and 20 trees from either the single dominant-codominant layer (CA) or the open canopy (SJ) were randomly selected. Two cores of 5 mm of diameter were taken from each tree at breast heigth (1.3 m) using an increment borer. Diameter at breast height (DBH) and total height were measured for each tree. In addition, stand competition of target trees was assessed by recording distance, azimuth, DBH, species and total height of all neighboring living trees with DBH > 7.5 cm within a circular plot of 10 m radius. Several competition indices were calculated: the distance independent indices *density* ($\mathrm{trees \cdot ha^{-1}}$), and *basal area* (BA, $\mathrm{m^{2} \cdot ha^{-1}}$); and the distance dependent index size ratio proportional to distance (*srd*) as $\mathrm{srd} = \sum_{i=1}^{n} (dbh_j / dbh_i) \cdot \left[1/ (dist_{ij} + 1) \right ]$. Differences between sites for height, DBH and competition indices were analysed using non-parametric Kruskal-Wallis rank sum tests. When significant differences were observed, we run multiple comparisons using the Dunn's-test with Bonferroni adjustment to correct significance values.  

Tree cores were air dried, glued onto wooden mounts and sanded. Annual radial growth (ring width, RW) was measured with a measuring device coupled to a stereomicroscope, with an accuracy of 0.001 mm. Individual ring series were first visually and statistically cross-dated with TSAP software (Rinntech, Heidelberg, Germany), using the statistics Gleichläufigkeit (GLK), t-value and the crossdating index (CDI). Cross-dating validation was finally done using COFECHA software [@Holmes1983].

The growth trends were analyzed at different time scales. To study the response of growth to the inter-annual variability of climate (short-term response), pre-whitened residual chronologies (RWI) were used. They were calculated from ratios between raw growth measurements and individual cubic splines with a 50 % frequency cutoff at 30 years [@Fritts1976]. Tree-ring width series were standardized and detrended using `dplR` [@Bunn2010]. Mean residual site chronologies were obtained computing the biweight robust mean of all prewhitened growth indices for the trees of the same site [@Cook1990]. The statistical quality of each chronology was checked via the expressed population signal. A threshold value of EPS > 0.85 was used to determine the cutoff year of the time span that could be considered reliable.

The long-term growth response was analyzed using basal area increment (hereafter BAI, $\mathrm{cm^2 \cdot year^{-1}}$). Theoretically, BAI represents a more accurate indicator of growth than ring-width, since it removes variation in growth attributable to increasing stem circumference after 30-40 years of juvenile increasing growth [@Biondi2008]. Raw ring-widths and measured DBH were used to compute BAI by subtracting twice the annual ring width from the annual diameter, starting from the measured diameter outside the bark [@Piovesan2008]. We used the following equation:

$$\mathrm{BAI} = \pi(r^2_{t}-r^2_{t-1})$$
where $r$ is the radius of the tree and $t$ is the year of tree-ring formation. For each individual tree, a mean BAI serie was calculated. Then, mean site BAI chronologies were obtained by averaging individual tree BAI time series. 

Both RWI and BAI chronologies were compared to different climatic indices to analyse the low and the high-frequency response to climate of trees. 

### Climate and growth 

Climate data were obtained from the European Daily High-Resolution Observational Gridded Dataset (E-OBS v16) [@Haylock2008]. Monthly precipitation and minimum and maximum temperatures were obtained at a 0.25 x 0.25 º resolution for the 1950-2016 period. We selected grid cells covering each sampled sites. The Standardized Precipitation-Evapotranspiration Index (SPEI) with a temporal scale of 6 months was used to characterize the drought conditions for the period 1961-2014. 

Relationships between residual site chronologies and climatic variables were assessed ussing bootstrapped Pearson's correlations estimated using `treeclim` package [@Zang2015]. 

### Disturbance analyses 
Disturbance chronologies were built using tree-ring width to identify abrupt and sustained increases (release events from competition) or decreases (supressions) in radial growth [@Nowacki1997] as indirect estimates of possible disturbance events (e.g. logging, neighbor mortality) in the past. By using a temporal window of reasonable length this method is intended to filter out the response to short-term changes in temperature and precipitation and minimizes the long-term growth trends [@Nowacki1997]. Growth changes (GC) were calculated for the individual tree-ring series using a 10-year running window as either positive (PGC) or negative (NGC) growth changes:

$$\mathrm{\% GC} = \left [ \frac{\left ( M1 - M2 \right )}{M2}\right] \times 100 $$
where $M1$ is preceding 10-year median and $M2$ is subsequent 10-year median. Medians were used since they are more robust estimator of central tendency than means [@Rubino2004]. Site disturbance chronologies were constructed by averaging the individual disturbances series annually. To separate growth peaks produced by disturbance events and expressing stand-wise disturbances from those by climate, we considered a threshold of 50 % of GC and more than 50% of the individual trees displaying the same growth changes was considered a stand-wise disturbance. 

In addition, the forest and managment history of our sampling sites was inferred from a detailed analysis of historical land-use changes. For this purpose, an exhaustive review of historical documents was done to compile information about socio-economical activities affecting forest (Table S4). 

```{r eval=FALSE, echo=FALSE}
# : land-uses, forest managment actions, mining, wildfires, etc. Several documentary sources were reviewed: historical documents and maps [*e.g.* @Titos1990]; detailed mining reports [*e.g.* @Maestre1858]; official information about recent wildfires events and forest management practices [@Bonet2016obsnev_forest]; livestock farming [*e.g.* @MorenoLlorca2016]; traditional irrigation ditches [@RuizRuiz2017] and other studies reviewing the socioeconomic dynamics of forest of Sierra Nevada at different scales [@JimenezOlivencia2015; @MorenoLlorca2016].
```

## Assessing ecosystem and tree individual resilience to drought

To evaluate the effects of drougth events on ecosystem resilience (using greeennes data) and tree individual resilience (using BAI growth data) we used resilience indices proposed by @Lloret2011: resilience (*Rs*), resistance (*Rt*), and recovery (*Rc*). The Resistance index, estimated as the ratio between the performance during and before the disturbance ($Rt=Drought / PreDrought$), quantifies the severity of the impact of the disturbance in the year it occurred. The Recovery index, computes as the ratio between performance after and during disturbance ($Rc = PostDrought / Drought$), represents the ability to recover from disturbance relative to its severity. Finally, the Resilience index ($Rs = PostDrought / PreDrought$) is the capacity to reach pre-disturbance performance levels. 
The values of these indices were computed for tree growth (BAI) and greenness (EVI mean) during each drought event. 2005 and 2012 were considered as single drought events. The predrought and postdrought values of each target variable (*i.e.* BAI or EVI) were computed as the mean value during a period of three years before and after the drought event respectively. A period of three years was chosen because we found similar results comparing periods of two, three and four years (Figure S6) and this length was used in other studies [*e.g.* @Gazol2018]. To put in context in a longer-term the resilience observed for 2005 and 2012 drought events, as well as to test whether the resilience indices were ecologically meaningful, resilience metrics for BAI data were additionally computed for the most severe drought events since 1950 and compared them with drought severity (Figure 6, Table S3). 

## Statistical analysis 

For each of the three resilience indices studied, robust two-way ANOVAs were used to test for differences between drought events (2005 and 2012) and the studied oak populations (northern and southern aspects). These tests were used because original and log-transformed data did not follow the assumptions of normality and homogeneity of variance [@Wilcox2012]. Robust measures of central tendency (M-estimator based on Huber's Psi) were used since they were close to the mean value in all cases [@Wilcox2012]. When running the robust ANOVA test, data were boostrapped 3000 times and trimmed automatically to control the potential influence of outliers. Post-hoc differences were assessed pairwise using a similar boostrap test. All the robust ANOVA and post-hoc tests were carried out using the `WRS2` package. The level of significance was set to 0.05 and adjusted for multiple comparisons. 

# Results

## Time trends in vegetation greenness

The analysis of greenness time trends showed that EVI in 78.9 % of the pixels of *Quercus pyrenaica* forests experienced a positive trend for the 2000-2016 period. The lowest values of EVI standardized anomalies for the studied period were recorded during the 2005 drought, and the minimum EVI values were expressed in northern (dry) populations (Figure 2). In fact, a "browning" episode ($\mathrm{EVI_{sa}} < -1$) was observed for *Q. pyrenaica* forests during this drought event, whereas no changes in greenness were observed in response to the 2012 drought (Figure 3). 

## Analysis of radial growth trends and disturbances 

The trees of the southern population were older than those from the northern one. In addition trees from the southern population at high altitude were taller and their growth was significantly faster than that of trees from the other two sites (Table 1, Figure 4). Both growth and height of trees from the northern and the low-elevation southern population were similar (Figures 4 and S6). Competition was similar among all sites. Yet plot basal area was greatest in CA-High (Table 1), despite the highest individual BAI growth at that site (Figure 4). Only trees from the southern sites showed significant positive growth trends since the late 1970s (Figure 4), being more pronunced for the highest site (CA-High).

Drought events produced reduction of RWI for all sites (Figure S3), being particularly evident the negative effect of the 1995 drought, when the greatest reduction of tree-growth were recorded for all chronologies. The southern sites (CA-High and CA-Low) showed weaker reductions of tree-growth than the northern site (SJ), especially for 2005 and 2012 (Figure S3). Tree-growth reduction followed a moisture-gradient, being weaker for the wetter site (CA-High). Trees from northern site also suffered a great reduction during 1945-1946 drought event (Figure 5). 

The northern site (SJ) showed two major release events (GC > 50 % occurring in more than 50 % of sampled trees): the first during the 1940 decade (the most evindent) and the second in the period 1995-2000. These periods alternate with periods of supression for over 40 years. The two southern sites showed no release events except for CA-High at the beginning of the 1830s and no suppression events in the last 50 years.

## Resilience to drought events at the ecosystem and individual tree levels
The exploration of the resilience metrics of tree-growth for severe drought events since 1950 (*i.e.* for the share period among the three chronologies, excluding juvenile years, Table S3) revealed a positive relation between drought severity and Recovery (*Rc*), significant for all oak populations (Figure 6). A similar pattern was obtained for Resilience (*Rs*) but only significant for SJ site (Figure 6). Non-significant patterns were obtained if we exclude 1995 from the analysis (Figure S8). Trees showed the highest value of tree-growth resilience for 1995, the worst drought event in our study area (Table S3), and particularly SJ, where our results suggest a major release event after 1995 (Figure 5).

During the last two drought events, resilience metrics for greenness and tree-growth were significantly different bewteen drought events (Table 3). The 2005 drought event reduced greennes and growth more than that of 2012 (Tables S1 and S2) but the two metrics of resilience generally covaried in the same direction. We obtained significantly higher Resilience (*Rs*) and Resistance (*Rt*) values for the 2012, the most severe event, than for 2005 in both variables (Tables S1-S2; Figure 7). Recovery (*Rc*) showed contrasting pattern for EVI and tree-growth. Recovery values for greenness was higher for 2005 than for 2012 drought event (Table S1, Figure 7). 

Resilience metrics significantly varied between sites, except for Resilience (*Rs*) of tree-growth (p = 0.534; Table 3). The two southern populations showed lower values for Recovery (*Rc*) than the northern site both for greenness and tree-growth (Tables S1 and S2). In contrast, Resistance and Resilience values were significantly higher for southern sites than for northern ones (Tables S1 and S2).

## Tree-growth response to climate
Tree-growth exhibited a significant greater response to water availability than to temperatures, as it can be expected for a Mediterranean site and a species located closer to its xeric/dry distribution limit (rear-edge). Cumulative precipitation of the hydrological year was the climatic variable with the highest relationship with growth and influenced similarly and positively the tree-growth for both northern and southern populations (Figure 8a). Precipitation of previous december was also positively correlated with tree growth in the northern population and in the highest location of the southern population. Hydrological, Spring and Summer SPEI showed a strong positive correlation with tree-growth (Figure 8a), being this relationship higher in the northern population (r > 0.6 *vs.* r < 0.5). Relationships with temperatures were lower than those with moisture variables. Yet there were differences between northern and southern populations: spring maximum temperature was the most significant limitation for the tree-growth of the southern population at both elevations while no significant limitations were recorded for northen site regarding the temperatures (Figure 8b and 8c). Remarkably minimum and maximum temperatures of current September positively influenced the tree-growth only in the northern population. No positive relationship with temperature was found for southern trees. 

# Discussion
## Rear-edge oaks show high resilience in response both to recent drought events and long-term climatic variability. 
Severe drought events negatively affected to primary and secondary growth of *Q. pyrenaica* forests in the rear edge of their distribution. Our findings show that 2005 and 2012 drought events, provoked reductions in greenness and tree-growth of these forests. These results are consistent with previous studies that reported significant reductions of tree-growth for this oak species during extreme drought events [@Corcuera2006; @GeaIzquierdo2014]. In fact, when we explore the effects of drought at a longer-term scale, we observed the greatest reduction of tree-growth during the 1995 drought, a characteristic pointer-year that caused severe and extensive damage to the Mediterranean vegetation across the Iberian Peninsula [@Penuelas2001; @Gazol2018]. 
Although the south of Iberian Peninsula suffered severe drought events in the last decades [*e.g* 2005 and 2012 were considered two of the worst recorded drought; @GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of *Q. pyrenaica* (78.9 % of the pixels showed a positive trend during the 2000 to 2016 period). This confirms previous findings that pointed out an increase in primary productivity for these forests [@PerezLuque2015onto], being the most productive ecosystems in this mountain region during that period [@Alcaraz2016obsnev_ndvi]. For tree-growth, we also observed positive trends in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4). Similar long-term trends were described for this species along its distribution range [@GeaIzquierdo2014] and could be related to the rising of the temperatures in the last decades [@GeaIzquierdo2014] particularly after 1970. Howewer, the observed pattern for our sites (a rear-edge) differs from the decline trends observed for several Mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P. sylvestris*; @SanchezSalguero2012; @Camarero2015; *Fagus sylvatica*; @Dorado2017]. 

Resilience (*Rs*) values observed in our study sites for the latest drought events in both tree-growth and greenness were close to 1, except for the tree-growth during 2005 event (Tables S1 and S2). Despite the 2012 drought event was more severe and intense than 2005, according with our analysis with SPEI data (Table S3), resilience values for greenness and tree-growth were greater for 2012 than for 2005. These results could be explained by the different timing of the two droughts, *i.e.* the 2012 drought was a winter drought [@Trigo2013] that occurred earlier than the 2005 drought. The latter lasted less than 2012 drought, but matched the period of maximum growth for *Q. pyrenaica* forests (Figure S7). Our results are in line with recently studies that indicate the timing of the drought as a key factor determining tree recovery after drought [@Huang2018], that also has been observed for other Mediterranean oak species [*e.g.* Holm oak; @Camarero2015b]. 

As we mentioned previously, we found strong declines of tree-growth during the most severe drought events occurred (1995 and 1999) (Table S3; Figure S3), matching similar declining patterns observed for other species in their rear edge [@SanchezSalguero2012; @Camarero2015]. But, interestingly, we obtained the highest values of resilience for tree-growth for these two drought events (Figure 6). In addition, the analysis from a long-term perspective of the severe droughts suggests a positive relation between the tree-growth resilience and drought severity (Figures 6). Surprisingly this result seems contradict our initial hypothesis in which we predict low resilience values for this oak species in their rear-edge after extreme drought events.  

Rear-edge populations live in environmental narrow margins, and small variations in environmental conditions can increase the vulnerability of the species [@Hampe2005]. A high vulnerability to drought is often assumed for populatios located at their rear-edge [@MartinezVilalta2018]. However we obtained positive trends in greenness and tree-growth for the rear-edge of *Q. pyrenaica* and high values of resilience to severe droughts were also reported. Our findings are in agreenment with those studies that have shown that the assumed higher vulnerability of dry edges does not necessarily hold [@Cavin2017]. @MartinezVilalta2018 pointed out the importance of the local adaptation and plasticity, and also of the local environmental factors on the vulnerability showed by rear-edge populations. The high values of resilience to drought reported here together with previously works showing high values of genetic resilience for those forests at Sierra Nevada, seem to be indicative of high local adaptation of this oak to this mountain range. In addition the local environmental conditions of the sites where this species inhabit at Sierra Nevada would explain the low vulnerability to drought showed by this oak. A key consideration for rear-edge populations is the different ways in which marginality can be defined, *i.e.* if it is defined according to geographic, climatic criteria, or according to other ecological factors [@MartinezVilalta2018]. In this sense, the high values of resilience to the drought events that we have observed, could suggest that the populations of *Q. pyrenaica* in Sierra Nevada are located in a geographical but not climatic rear-edge. 

A remarkably result is the higher values of resilience for tree-growth obtained during the worst and more severe drought events in our study area (1995 and 1999) (Table S3; Figure 6). A plausible explanation could be the water availability previously and after the drought event. For instance, cumulative precipitation of the 1992-1994 years (pre-1995 drought) had precipitation values below the mean of the whole period (Figure S1), while 1996-1998 years (post-1995 drought) showed a wet opposite pattern (*e.g.* first months of 1996 recorded the highest cumulative rainfall of the past century for our southern sites). This could explain the higher values of recovery found for 1995 (Figure 6). The high resilience values obtanied for the 1995 drought in the most xeric site (SJ) could be related with drought-induced mortality occurred in this area (*personal communications*). 

Greenness and tree-growth of *Q. pyrenaica* showed different sensitivity to severe droughts. During the last drought events EVI was reduced to 90.25 % (Table 1), whereas BAI was reduced to 70.83 % (Table S2), suggesting a lower sensitivity of the greenness than tree-growth to drought, particularly for sites under drier climate. For our driest site (SJ, the northern one), BAI were reduced to 44.5 % respect to that of the preceding period during 2005 drought event, whereas the EVI was reduced to 81.9 % (Tables S1 and S2). These findings are in accordance with previously works which showed that tree-growth are more sensitive metrics of forest resilience than net primary productivty [*e.g.* @Babst2013; @Coulthard2017; @Gazol2018]. Besides the different scales of both approaches, *i.e.* tree-based *versus* pixel-based; the growth reduction seems to be more mediated by sink that by source limitations [@Gazol2018]. Response of greenness to drought not only reflects response of tree, because remote sensing indices (EVI) captures signals from all vegetation covered by the pixel (not only tree but also understory vegetation) [see @Dionisioetal2012IPoGEC]. In this sense, recent results indicate that remote sensing models underestimate the impact of drought on primary production [@Stocker2019], as soil moisture is not taken into account by this satellite estimates. This is particularly important for *Q. pyrenaica* forests, as soil moisture is a key growth-related factor for this species [@GeaIzquierdo2014]. On the other hand, tree-ring records allow to contextualize remote sensing data by reflecying climate or disturbance induced tree-growth anomalies over decades to centuries [@Babst2017] and provide an accurate measure of growth responses to droughts [@Bhuyan2017; @Gazol2018]. Combined anlaysis, such the approach used in our study, may also allow climate-induced variability in forest growth to be disentangled from that driven by community-level ecological processes [@Babst2017], since spatio temporal patterns of tree-ring measurement and earth observations (*i.e.* EVI) reflects processes at different spatio-temporal scales [Babst2018].  

## Microenvironment shapes differential sensitivity to climate and drought of rear-edge oaks 

Our results showed differences for greenness and tree-growth between northern and southern oak populations (Table 3). The former, which have drier conditions (Table 1), were more impacted by the drought events than the latter. For instance, the northern site showed more negative EVI standardized anomalies (higher browning intensity) than the southern sites during the 2005 drought event (Figure 2). In addition, the stronger correlations of tree-growth with SPEI (Hydrological and summer) observed for northern site (Figure 8), can be interpreted as higher sensitivity to drought of a drier site [@GeaIzquierdo2014]. 

Moisture availability has been reported as the most limiting factor driving radial growth of *Q. pyrenaica* along their distribution range in Iberian Peninsula [@GeaIzquierdo2014]. We found a singnificant positive relation between precipitacion (hydrological year and previous December) and tree-growth (Figure 8a). Our results are consistent with previous studies highlighting the influence of precipitation on tree-ring growth in oaks [*e.g.* @GeaIzquierdo2014; @GonzalezGonzalez2014; @GarciaGonzalez2017]

A striking result is the difference for tree growth between sites (Figure 4). The trees of CA-High site, which are located around 1900 m. *a.s.l.* and representing the upper limit of the treeline of the species in this southernmost location, showed higher BAI than those located at low-elevations sites (CA-Low and SJ) (Figure 4). This is especially interesting for southern sites, which are very close to each other. Our findings are in line with previous works that pointed out that tree growth and tree responses to drought are site-dependent (*e.g.* soil features, tree competence, etc.) [@Babst2013; @VicenteSerrano2014b], particularly for rear-edge populations [@Cavin2017; @Dorado2017b]. The higher values of tree growth that we observed in CA-High site could be related with lower water stress at high-elevation sites, since moisture availability is a key factor limiting tree growth for *Q. pyrenaica* [@GeaIzquierdo2014]. In fact, soil analyses along an elevational gradient in this location revealed higher content of available water for forest soils located on high elevation [@CoboDiaz2017]. Recent works have found a clearly effect of elevation on the resistance to drought of several tree-species, where stands located at higher elevations were less drought sensitive.  [@Bhuyan2017]. 

```{r eval=FALSE, echo=FALSE}
# In this sense, a remarkable observation is the presence of traditional irrigation ditches which could provide an additional water supply. In most of the high-mountain watersheds of Sierra Nevada, there are systems of historical irrigation channels, know as *acequias de careo*, that was used since Middle Age to cultivated these valleys [@MartinCivantos2014]. These ditches run through the hillsides of valleys releasing water through several points that filtrate the water to recharge the aquifer and then irrigates the foot of the slopes [@MartinCivantos2014; @Jodar2017]. Several works have recently described the network of traditional ditches for our southern sites (Cáñar), with detailed information about water use [@RuizRuiz2017] and its hydrological functioning [@MartinMontanes2015]. There is a ditch (*Acequia de la Era Alta*) located uphill the CA-High site, which functions from March to June [@RuizRuiz2017]. It brings water from snowmelt and from a nearby stream, soaking the *Q. pyrenaica* forests. This traditional system could supply an extra of water that could be used by the trees located downstream of these channels. This extra of water is particularly important for trees of this dry-rear edge and could explain the higher values of BAI for trees located in CA-High site. 
``` 

In addition to the lower water stress, it is likely that the sites at higher elevation had a lower anthropic impact and the abandonment of traditional uses was earlier, due to their farthest location from human settlements and therefore less access to forest resources (Table S4). 

## Land-use legacies shape distribution and sensitivity to climate change at the read-edge

The Mediterranean mountains, even at high elevations, have suffered strongly transformation of the landscapes driven by human activities throughout history. Multiple evidence from palaeoecological studies indicated an intense human impact on vegetation of Sierra Nevada since 3000 cal year BP [@GarciaAlix2017]. From that moment onwards, increases on fire occurrence and both grazing and mining activities, were recorded for this mountain region, with an sharp intensification of the human activities in the last 150 years [@GarciaAlix2017]. Mining, forest clearings, fuelwood and charcoal exploitations, pastoralism and also wars, have strongly impacted on the forest resources in Sierra Nevada, and on Pyrenean oak forest, as can be derived from the results of the review of historical documents (Table S4). As a consequence of all those activities a loss of about 90% of broadleaf *Quercus* species cover from medium and low elevation occurred in this mountain region [@JimenezOlivencia2015]. 


 

Our study sites have been subjected to an intense exploitation of forest resources in history and landscape transformation. Although the estimated age for our sampled trees was not more than 180 years (Tables 1, 2), several documents reported the presence of oaks before that date. For instance, the inventories of trees made by the Spanish Navy during the second half of 18th century, reported for Cáñar and SJ sites more than two millions and 700 000 trees respectivley, most of them juveniles, suggesting recent wood fellings [@Cruz1991]. These numbers are a reflect of the intense exploitation of the forest resources occurring up to the end of nineteenth century [@Calatrava2019] and also concur with the regeneration peaks observed for several Iberian oak woodlands [e.g. @GeaIzquierdo2014; @Gea-Izquierdo2015; @Dorado2017].  

We found similar tree competence levels in our sampled sites (Table 1), but differences in tree size and age suggest different management origin (*i.e.* land-use legacy). The analysis of growth changes revealed differences in forest history between sites (Figures 5 and S4), driven mainly by the differential pattern of the natural resource use [@JimenezOlivencia2015] (Table S4). On the one hand, the northern sites (*e.g.* San Juan), show a sequentially distribution of the land uses along the elevational gradient [@JimenezOlivencia2015]: grasslands and shrublands for cattle farming are located at high elevations; then forests formation with some croplands; and at lowlands, irrigated terraces with tree crops. Other activities (*e.g.* mining) could also determine the use of natural resources, and therefore the forest structure. For instance, the SJ oak woodland is located in an area with a high concentration of mines and quarries that have been exploited intermitently throughout history (Table S4). On the other hand, woodland areas of the southern slopes (Cáñar) are mixed with a greater percentage of croplands, even reached high elevation [@Calatrava2019]. Forest resources, like firewood, charcoal and acorns, have been continuously exploited in southern sites through history [@Valbuena2013]. In Cáñar site, there were an arraigated tradition of charcoal extraction since fifteenth century [@JimenezSerrano2004] up to middle of the 20th century. From this moment to the present, there were a sharp decrease of the wood extraction, mainly due to rural abandonment. Acorn resources have been tradiotionally exploited in this location up to 1950s (Tabla S4). 

Those different patterns of resource uses can help to understand the tree-growth changes recorded in our chronologies (Figure 5). We observed a tree-growth release event at 1940s in the SJ oak woodland (Figures 5, S4) which concurs with one period of maximum mining activity for this area (1925 to 1957) (Table S4). During this period there were an increase on the use of timbers for the tunnels of the mines and several furnaces that required great quantities of fuelwood to melt the mineral were active in this area [@Titos1990]. This heavily exploitation of the forest resources could affect to a major part of this oak woodland, since percentage of trees affected by GC > 50 % reaches values above 50 % (Figure S4). Our results also coincide with a concurrent but less intense tree-growth release event reported for a closed oak woodland [@GeaIzquierdo2014]. In addition, paleoecological studies carried out in nearby alpine bogs have recorded increases in the heavy metals concentration since the end of 18th century until mid-20th, which coincides with the maximum activity of the mining in this area [@GarciaAlix2017]. The other release event observed for SJ site during 1995-2000 was lower than the occurred at 1940, but affecting more trees (Figures 5, S4). We revised the forest practices carried out in this area in the last 30 years [@Bonet2016obsnev_forest], but we did not find any clearing or cutting event during 1995 - 2000 period. Two non-mutually exclusive ways could explain the tree-growth release observed. Firstly it could be related with a natural drought-induced mortality event after 1995 (*personal communications*) as was reported for Mediterranean tree species [*e.g.* @Penuelas2000; @Lloret2004]. On the other hand, we obtained strong positive correlations of SPEI with tree-growth for this site (Figure 8), which suggests a high sensitivity to water availability of this slightly more xeric site (Table 1) [@GeaIzquierdo2014]. Previous works reported a sharp decline (or non-production) in latewood production during extreme drought events [@Corcuera2006]. Since latewood are less vulnerable to embolism than earlywood vessels [@Corcuera2006], the reduction or even the non-production of former could negatively affect to tree-growth and also enhanced the mortality, particularly for drier sites [@Corcuera2006; @GeaIzquierdo2014].

For southern sites, the only positive release events showed for CA-High site could be related with conversion from closed forest to an open silvopastoral system, a common management applied in the past [@Canellas2004; @GeaIzquierdo2014] documented in this site [@Valbuena2013].  

Climatic drivers are key factors determining the growth of tree species, especially at the rear edge of their distribution but in Mediterranean forests we must also considered the management history, *i.e* *land-use legacies* [@Navarro2013; @DoblasMiranda2017; @Penuelas2017] since, as we inferred from our results, the land-use legacy can conditioned tree-growth and resilience of tree species. Although the abandonment of the traditional activities have provoked a stagnation for this species [@Canellas2004], we have observed an increase in the growth, especially in the high elevation sites, where the water is less-limiting resource. This pattern concurs with a forest expansion into marginal abandonded croplands and an increase in the tree-cover [@CamachoOlmedo2002].  

# Conclussions 
Severe drought events have provoked important reductions in primary and secondary growth of *Q. pyrenaica* forests at Sierra Nevada mountain range. However, despite the expected vulnerability for those relict stands, we observed positive trends of both primary (*i.e.* greenness) at ecosystem scale and secondary growth at tree level in our study site, particularly at the most humid site along the climatic gradient analyzed. Surprisingly we observed high resilience values to drought, particularly at long-term scales. These results are consistent with research showing high genetic resilience values and overall suggest this mountain region still acting as refuge for deciduous *Quercus*. The differences observed in tree-growth and resilience to drought between very close sites show that responses to drought were site-dependent. This is particularly relevant for rear-edge populations where topographic and biophysical variablity facilitates the existence of microrefugia, as reflected along the climatic gradient analysed [@VilaCabrera2019]. The analysis of tree-growth dynamics revealed suppression and release events that are consistent with legacies left by land-use in local forest dynamics, as inferred from an exhaustive review of historical documents. In this sense our results highlight the importance of land-use legacies for highly transformed Mediterranean systems. This is relevant for tree species with higher sensitivity to climate change, such us *Q. pyrenaica*, not only for conservation per se of this species, but for all ecosystem services that these singular forests offer at Mediterranean mountain. Following our study it is necessary to disentangle how ecosystem services and stand dynamics are affected by the combination of climate change and land-use legacies in highly transformed landscapes like the one studied here.

# Acknowledgments

We would like to thank Sammy Lucy Behle, M. Suarez-Muñoz and F.J. Bonet-García for their assistence in the field monitoring. A. Reyes and J. Blanco helped in the filtering of satellite data. F.J. Cano-Manuel and F.J. Navarro provided worthwhile information about oak management projects in Sierra Nevada. We also are very thankful to F.J. Bonet-García for his valuable comments in an earlier version of the manuscript. Environment Department of Andalusian Regional Government and Director of Sierra Nevada National and Natural Park provided permission for field work. This research work was conducted in the collaborative framework of the “Sierra Nevada Global Change Observatory” Project from the Environment Department of Andalusian Regional Government (with the support of European Union FEDER Project). We also thank to LIFE-ADAPTAMED (LIFE14 CCA/ES/000612): *Protection of key ecosystem services by adaptive management of Climate Change endangered Mediterranean socioecosystems* and H2020 project *ECOPOTENTIAL: Improving future ecosystem benefits through earth observations* (http://www.ecopotential-project.eu/) (No 641762), for their funding support. 

# References 
