---
output:  
  word_document:
    reference_docx: "../../man/template_tables_portrait.docx"
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, echo=FALSE}
library(tidyverse)
library(here)
library(pander)
library(kableExtra)
library(knitr)
library(dplyr)
```


**Table 2.** Characteristics of the mean tree ring chronologies. Values of the length year in parenthesis indicate years replicated more than five series. RW = mean annual ring width (standard deviation in parenthesis). MS = mean sensitivity. AR(1) = mean autocorrelation of raw series. Rbt = mean correlation between series. EPS = mean expressed population signal. EPS and Rbt are calculated for the mean residual chronologies of growth indices. 

```{r, echo=FALSE}
topo <- read.csv(here("/data/sites_summary", "site_general.csv"), header = TRUE)
ages <- read.csv(here("/data/dendro_summary", "site3_avg_ages.csv"), header = TRUE)
dendro_summ <- read.csv(here("/data/dendro_summary", "site3_dendro_summary.csv"), header = TRUE)
biotic <- read.csv(here("/data/competence", "biotic_stand_features.csv"), header = TRUE)

t1raw <- topo %>% inner_join(ages, by='site') %>% 
  inner_join(dendro_summ, by='site') %>% 
  inner_join(biotic, by='site')


t1raw <- t1raw %>% mutate(name = case_when(
  site == 'caH' ~ 'C치침ar-High', 
  site == 'caL' ~ 'C치침ar-Low',
  site == 'sj' ~ 'San Juan')) 


t2 <- t1raw %>% dplyr::select(
  site, first_year, last_year, 
  length_year, length_year5, 
  n_trees, n_radii,
  rw_mean, rw_std, ms2, ar1, rbar.wt, eps)  %>% 
  mutate(pare = ")") %>% 
  mutate_at(c(8:13), round, 3) %>% 
  unite(length_year, c(length_year,length_year5), sep=" (") %>%
  unite(RW, c(rw_mean,rw_std), sep=" (") %>%
  mutate(
    length_year = paste0(length_year, pare),
    RW = paste0(RW, pare)) %>% 
  dplyr::select(-pare)
  
  
colnames(t2) <- c("Site", "First year", "Last year", "Length (years)",  
                   "# trees", '# cores', "RW (mm)", "MS", "AR(1)", "Rbt", "EPS") 

pander(t2,
       split.table = Inf)

```



