---
title: "disturbance chronologies"
author: "AJ Perez-Luque (@ajpelu)"
date: "2017 Feb"
output:  
  md_document:
    variant: markdown_github
bibliography: references.bib 
csl: ecology.csl
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8,warning=FALSE, message=FALSE)
```


```{r, warning=FALSE, message=FALSE}
library("tidyverse")
library("stringr")
library("dplR")
library("knitr")
# library("detrendeR")
library("pander")
library('boot')
library('gtable')
library('grid')
```

# Disturbance chronologies 

* Vamos a calcular las disturbance chronologies siguiendo la aproximación de [@Nowacki1997](varios ejemplos se pueden ver en [@GeaIzquierdo2014, @Dorado2017])

* Utilizaremos tres chronologias: SJ, CAH y CAL (ver [./analysis/analysis_chronologies.md]('./analysis/analysis_chronologies.md)) 


## Prepare data 

* Leer datos `rwl` de SJ y CA
* Leer datos de diametros de los focal tree

```{r, echo=FALSE}
machine <- 'ajpelu'
#machine <- 'ajpeluLap'
di <- paste0('/Users/', machine, '/Dropbox/phd/phd_repos/qpyr_dendro/', sep='')

# sj 
sj <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_sanjuan/sn_sanjuan.rwl'), format="tucson")

# canar 
ca <- read.rwl(fname=paste0(di, '/data_raw/dendro_ring/sn_canar/sn_canar.rwl'), format="tucson")

# Read diameters data
compete <- read.csv(file=paste0(di, '/data_raw/dendro_competence.csv'), header=TRUE, sep=',')
```

```{r}
source(paste0(di, 'script/R/rw_byTree.R'))
source(paste0(di, 'script/R/bai_piovesan.R'))
source(paste0(di, 'script/R/computeGC.R'))
```


* Crear dataframes `rwl` por cada sitio CA_High, CA_Low, SJ_High. SJ_Low

    * :red_circle: ojo he cambiado los nombres de las series en SJ y en CA
    
```{r}
# Replace SNA by SJ and SNB by CA
names(ca) <- stringr::str_replace(names(ca), "SNB", "CA") 
names(sj) <- stringr::str_replace(names(sj), "SNA", "SJ")

# Remove g in name of some cores of CA. 
names(ca) <- stringr::str_replace(names(ca), "g", "")
```

```{r}
# Create subset to compare between sites 
caL <- ca[,c("CA0101","CA0102","CA0201","CA0202","CA0301","CA0302","CA0401","CA0402","CA0501","CA0502",
             "CA0601","CA0602","CA0701","CA0702","CA0801","CA0802","CA0901","CA0902","CA1001","CA1002",
             "CA2601","CA2602","CA2701","CA2702","CA2801","CA2802","CA2901","CA2902","CA3001","CA3002")] 
caH <- ca[, c("CA1101","CA1102","CA1201","CA1202","CA1301","CA1302","CA1401","CA1402","CA1501","CA1502",
              "CA1601","CA1602","CA1701","CA1702","CA1801","CA1802","CA1901","CA1902","CA2001","CA2002",
              "CA2101","CA2102","CA2201","CA2202","CA2301","CA2302","CA2401","CA2402","CA2501","CA2502")]

```

* Lectura y preparación de datos de diámetro

```{r} 
# Prepare Diameter data 

# Compute diameter (mm)
compete <- compete %>% 
  mutate(dn_mm = (perim_mm / pi))

# Change name focal according to loc
compete <- compete %>% 
  mutate(id_focalLoc = stringr::str_replace_all(id_focal, c("A" = "SJ", "B" = "CA")))

         
# Get only focal trees, and only selected variables 
ft <- compete %>% 
  filter(sp=='Focal') %>% 
  filter(id_focal!='Fresno') %>% 
  dplyr::select(id_focal, id_focalLoc, loc, dn_mm, height_cm) 

# Set levels of eleveation 
ca_lowcode <- c(paste0('CA', str_pad(1:10, 2, pad='0')),
            paste0('CA', 26:30))
ca_highcode <- paste0('CA', 11:25)

ft <- ft %>% 
  mutate(site = as.factor(
    ifelse(id_focalLoc %in% ca_lowcode, 'CAL', 
           ifelse(id_focalLoc %in% ca_highcode, 'CAH', 'SJ'))))
```


# Summary dendrochronology statistics

* Por cada site (CAH, CAL, SJ) calculamos los siguientes datos: 
    * age: mean + sd (average of all series)
    * year: min, max of all series
    * year range
    * autocorrelation order 1 (ar1): mean + sd 
    * seires intercorrelation (icor): mean + sd 
    
## By site 
```{r}
objects_rwl <- c('caL','caH','sj')

out <- c() 

for (i in objects_rwl){ 
  
  aux <- get(i)
  # Stats of singles RWL
  aux_stats <- rwl.stats(aux)
  # Add site 
  aux_stats$site <- as.factor(i)
  
  # Compute interseries correlation 
  aux_stats$interser_cor_val  <- interseries.cor(aux)$res.cor
  aux_stats$interser_cor_p  <- interseries.cor(aux)$p.val
  
  out <- rbind(out, aux_stats)
  }


rwl_by_site <- out %>% group_by(site) %>% 
  summarize(age_mean = round(mean(year),2), 
            age_sd = round(sd(year), 2),
            year_min = min(first),
            year_max = max(last), 
            year_range = year_max - year_min + 1, 
            ar1_mean = round(mean(ar1), 3),
            ar1_sd = round(sd(ar1), 3),
            icor_mean = round(mean(interser_cor_val), 3),
            icor_sd = round(sd(interser_cor_val), 3)) %>% 
  as.data.frame() 

rwl_by_site %>% 
  write.csv(file=paste(di, "data/proto_tables/site_3_dendro.csv", sep=""), row.names = FALSE)

pander(rwl_by_site, caption='Dendrochronological summary by sites (3)') 
```



# Aggregate RW by tree 

* Agregar valores medios de RW por site (obtenemos sj_tree / caL_tree, caH_tree) 
* ver fun rw_byTree o utilizar treeMean (dplR)

```{r}
# Remember snc = structure of core name SJ0101 (site | tree | core)
sj_tree <- rw_byTree(sj, snc =c(2,2,2), locname = 'SJ')
caL_tree <- rw_byTree(caL, snc =c(2,2,2), locname = 'CA')
caH_tree <- rw_byTree(caH, snc =c(2,2,2), locname = 'CA')

# Si decido usar treeMean debería ser (solo un ejemplo)
# ids_sj <- read.ids(sj, stc = c(2, 2, 2))
# sj_meantree <- treeMean(sj, ids = ids_sj,  na.rm=TRUE)
```

* Crear diferentes dataset de diametro por sitio 
```{r}
diam <- ft %>%
  mutate(diameter = dn_mm, 
         id = id_focalLoc) %>%
  dplyr::select(id, diameter, site) %>% 
  split(.$site) 


d_caH <- diam$CAH[,c('id','diameter')]
d_caL <- diam$CAL[,c('id','diameter')]
d_sj <- diam$SJ[,c('id','diameter')]
```

# Disturbance chronologies 

* calcular chronologies con media y mediana. 
* ver [./analysis/analysis_disturbance_chronos.md](/analysis/analysis_disturbance_chronos.md))


```{r}
gc_caL <- computeGC(caL_tree, ws=10, prefijo = 'CA')
gc_caH <- computeGC(caH_tree, ws=10, prefijo = 'CA')
gc_sj <- computeGC(sj_tree, ws=10, prefijo = 'SJ')
```

### Plot (mean) 
```{r}
# Plot mean value by year 
# dataframe for plot 
gc_todos <- c() 

datasetes <- c('gc_caL', 'gc_caH', 'gc_sj')

for (d in datasetes){ 
  # get df
  df <- get(d)
  
  aux_df <- df %>%
    group_by(year) %>% 
    summarise(ntrees = length(ntrees),
              gc_m = mean(gc_mean),
              gc_m_sd = sd(gc_mean),
              gc_m_se = sd(gc_mean) / sqrt(n()), 
              gc_md = mean(gc_median),
              gc_md_sd = sd(gc_median),
              gc_md_se = sd(gc_median) / sqrt(n())) %>%  
    mutate(site=paste(d))
  
  gc_todos <- rbind(gc_todos, aux_df)
}


gc_todos %>% ggplot(aes(x=year, y=gc_m)) + 
  geom_bar(stat='identity', fill='gray', colour='gray') + theme_bw() + 
  # geom_point() +
  geom_errorbar(aes(ymin = gc_m - gc_m_se, ymax = gc_m + gc_m_se), colour='gray') +
  facet_wrap(~site, nrow = 3, scales = 'free_y',
             labeller=as_labeller(c('gc_caH' = 'Canar High', 
                                 'gc_caL' = 'Canar Low',
                                 'gc_sj' = 'San Juan'))) +
  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue') +
  geom_hline(yintercept=c(50, -50), linetype=2, colour='red') + 
  ylab('% GC (mean)') + 
  theme(panel.grid.minor = element_blank(), 
        strip.background = element_rect(colour='black', 
                                        fill='white'))
```

### Plot (median) 
```{r}
gc_todos %>% ggplot(aes(x=year, y=gc_md)) + 
  geom_bar(stat='identity', fill='gray', colour='gray') + theme_bw() + 
  # geom_point() +
  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray') +
  facet_wrap(~site, nrow = 3, scales = 'free_y',
             labeller=as_labeller(c('gc_caH' = 'Canar High', 
                                 'gc_caL' = 'Canar Low',
                                 'gc_sj' = 'San Juan'))) +
  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue') +
  geom_hline(yintercept=c(50, -50), linetype=2, colour='red') + 
  ylab('% GC (median)') + 
  theme(panel.grid.minor = element_blank(), 
        strip.background = element_rect(colour='black', 
                                        fill='white'))
```

### Plot (correcting the scale) (mean)
```{r}
# Create a transformation scale 
# https://stackoverflow.com/questions/37446064/i-need-ggplot-scale-x-log10-to-give-me-both-negative-and-positive-numbers-as-o 

pos_neg_log <- scales::trans_new('signed_log', 
                                 transform = function(x) sign(x)*log(abs(x)),
                                 inverse=function(x) sign(x)*exp(abs(x)))


# solution for arrange
# http://felixfan.github.io/stacking-plots-same-x/

                 
a <- gc_todos %>% 
  filter(site != 'gc_sj') %>% 
  ggplot(aes(x=year, y=gc_m)) + 
  geom_bar(stat='identity', fill='gray', colour='gray') + theme_bw() + 
  geom_errorbar(aes(ymin = gc_m - gc_m_se, ymax = gc_m + gc_m_se), colour='gray', width=.3, size=.2) +
  # scale_y_continuous(trans = pos_neg_log) +
  facet_wrap(~site, nrow = 2,
             labeller=as_labeller(c('gc_caH' = 'Canar High', 
                                 'gc_caL' = 'Canar Low'))) +
                                # 'gc_sj' = 'San Juan'))) + 
  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue', alpha=.4) +
  geom_hline(yintercept=c(50, -50), linetype=2, colour='red', alpha=.4) + 
  ylab('% GC (mean)') + 
  theme(panel.grid.minor = element_blank(), 
        strip.background = element_rect(colour='black', 
                                        fill='white'),
        axis.title.x = element_blank(),
        axis.text.x = element_blank()) 
  


b <- gc_todos %>% 
  filter(site == 'gc_sj') %>% 
  ggplot(aes(x=year, y=gc_m)) + 
  xlim(layer_scales(a)$x$range$range[1],layer_scales(a)$x$range$range[2]) +
  geom_bar(stat='identity', fill='gray', colour='gray') + theme_bw() + 
  geom_errorbar(aes(ymin = gc_m - gc_m_se, ymax = gc_m + gc_m_se), colour='gray', width=.3, size=.2) +
  scale_y_continuous(trans = pos_neg_log, 
                     breaks=c(-1000, -100, 0, 100, 500)) +
  facet_wrap(~site, nrow = 2,
             labeller=as_labeller(c('gc_sj' = 'San Juan'))) + 
  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue', alpha=.4) +
  geom_hline(yintercept=c(50, -50), linetype=2, colour='red', alpha=.4) + 
  ylab('') + 
  theme(panel.grid.minor = element_blank(), 
        strip.background = element_rect(colour='black', 
                                        fill='white')) 


grid.newpage()
grid.draw(rbind(ggplotGrob(a), ggplotGrob(b), size='last'))
```

### Plot (correcting the scale) (median)
```{r}
c <- gc_todos %>% 
  filter(site != 'gc_sj') %>% 
  ggplot(aes(x=year, y=gc_md)) + 
  geom_bar(stat='identity', fill='gray', colour='gray') + theme_bw() + 
  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) +
  # scale_y_continuous(trans = pos_neg_log) +
  facet_wrap(~site, nrow = 2,
             labeller=as_labeller(c('gc_caH' = 'Canar High', 
                                 'gc_caL' = 'Canar Low'))) +
                                # 'gc_sj' = 'San Juan'))) + 
  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue', alpha=.4) +
  geom_hline(yintercept=c(50, -50), linetype=2, colour='red', alpha=.4) + 
  ylab('% GC (median)') + 
  theme(panel.grid.minor = element_blank(), 
        strip.background = element_rect(colour='black', 
                                        fill='white'),
        axis.title.x = element_blank(),
        axis.text.x = element_blank()) 
  


d <- gc_todos %>% 
  filter(site == 'gc_sj') %>% 
  ggplot(aes(x=year, y=gc_md)) + 
  xlim(layer_scales(a)$x$range$range[1],layer_scales(a)$x$range$range[2]) +
  geom_bar(stat='identity', fill='gray', colour='gray') + theme_bw() + 
  geom_errorbar(aes(ymin = gc_md - gc_md_se, ymax = gc_md + gc_md_se), colour='gray', width=.3, size=.2) +
  scale_y_continuous(trans = pos_neg_log, 
                     breaks=c(-1000, -100, 0, 100, 500)) +
  facet_wrap(~site, nrow = 2,
             labeller=as_labeller(c('gc_sj' = 'San Juan'))) + 
  geom_hline(yintercept=c(25, -25), linetype=2, colour='blue', alpha=.4) +
  geom_hline(yintercept=c(50, -50), linetype=2, colour='red', alpha=.4) + 
  ylab('') + 
  theme(panel.grid.minor = element_blank(), 
        strip.background = element_rect(colour='black', 
                                        fill='white')) 


grid.newpage()
grid.draw(rbind(ggplotGrob(c), ggplotGrob(d), size='last'))
```


## Number of trees with %GC > 50% (mean)
```{r}

computeGCtrees <- function(gcdf, treshold){ 
  
  th <- as.numeric(treshold)
  
  # Get total tree by year 
  n_tree <- gcdf %>% group_by(year) %>% summarise(total_tree = length(ntrees)) 
  
  # Get n tree with GC positve upper treshold 
  n_pos <- gcdf %>% dplyr::filter(gc_mean > th) %>% group_by(year) %>% summarise(n_pos = length(ntrees)) 
  
  # Get n tree with GC negative lower treshold 
  n_neg <- gcdf %>% dplyr::filter(gc_mean < (-1*th)) %>% group_by(year) %>% summarise(n_neg = length(ntrees)) 
  
  # Join and compute percentage 
  outgc <- n_tree %>% 
    left_join(n_pos, by='year') %>% 
    left_join(n_neg, by='year') %>% 
    mutate(per_pos = round((n_pos / total_tree) * 100, 2),
         per_neg = round((n_neg / total_tree) * 100, 2))
  
  return(outgc)
}
  

gc_trees<- c()

datasetes <- c('gc_caL', 'gc_caH', 'gc_sj')

for (d in datasetes){ 
  # get df
  df <- get(d)
  
  au <- df %>% computeGCtrees(treshold = 50) %>% data.frame() %>% mutate(site = paste(d))
  gc_trees <- rbind(gc_trees, au)
} 

# secondary axis
# https://whatalnk.github.io/r-tips/ggplot2-secondary-y-axis.nb.html

# Plot 
gc_trees %>% ggplot() +
  geom_bar(aes(x=year, y=per_pos), fill='blue', colour='blue', stat = "identity", position = "dodge") + 
  geom_bar(aes(x=year, y=(-1)*per_neg), fill='red', colour='red', stat = "identity", position = "dodge") +
  facet_wrap(~site, nrow = 3, labeller=as_labeller(c('gc_caH' = 'Canar High', 
                                 'gc_caL' = 'Canar Low',
                                 'gc_sj' = 'San Juan'))) + 
  scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
  geom_line(aes(x=year, y=total_tree), colour = 'gray') + 
  theme_bw() + ylab('%trees GC > 50 %') +
  theme(panel.grid.minor = element_blank(), 
        strip.background = element_rect(colour='black', 
                                        fill='white')) 
  
```


## Number of trees with %GC > 50% (median)
```{r}

computeGCtrees <- function(gcdf, treshold){ 
  
  th <- as.numeric(treshold)
  
  # Get total tree by year 
  n_tree <- gcdf %>% group_by(year) %>% summarise(total_tree = length(ntrees)) 
  
  # Get n tree with GC positve upper treshold 
  n_pos <- gcdf %>% dplyr::filter(gc_median > th) %>% group_by(year) %>% summarise(n_pos = length(ntrees)) 
  
  # Get n tree with GC negative lower treshold 
  n_neg <- gcdf %>% dplyr::filter(gc_median < (-1*th)) %>% group_by(year) %>% summarise(n_neg = length(ntrees)) 
  
  # Join and compute percentage 
  outgc <- n_tree %>% 
    left_join(n_pos, by='year') %>% 
    left_join(n_neg, by='year') %>% 
    mutate(per_pos = round((n_pos / total_tree) * 100, 2),
         per_neg = round((n_neg / total_tree) * 100, 2))
  
  return(outgc)
}
  

gc_trees<- c()

datasetes <- c('gc_caL', 'gc_caH', 'gc_sj')

for (d in datasetes){ 
  # get df
  df <- get(d)
  
  au <- df %>% computeGCtrees(treshold = 50) %>% data.frame() %>% mutate(site = paste(d))
  gc_trees <- rbind(gc_trees, au)
} 

# secondary axis
# https://whatalnk.github.io/r-tips/ggplot2-secondary-y-axis.nb.html

# Plot 
gc_trees %>% ggplot() +
  geom_bar(aes(x=year, y=per_pos), fill='blue', colour='blue', stat = "identity", position = "dodge") + 
  geom_bar(aes(x=year, y=(-1)*per_neg), fill='red', colour='red', stat = "identity", position = "dodge") +
  facet_wrap(~site, nrow = 3, labeller=as_labeller(c('gc_caH' = 'Canar High', 
                                 'gc_caL' = 'Canar Low',
                                 'gc_sj' = 'San Juan'))) + 
  scale_y_continuous(sec.axis = sec_axis(~ ., name = '# trees', breaks = c(0,15,30))) +
  geom_line(aes(x=year, y=total_tree), colour = 'gray') + 
  theme_bw() + ylab('% trees GC > 50 % (median)') +
  theme(panel.grid.minor = element_blank(), 
        strip.background = element_rect(colour='black', 
                                        fill='white')) 
  
```

# TRADER approach 
Vamos a utilizar la aproximación del pkg `trader` [Altman2014]

## Nowacki & Abrams 1997
Mismo análisis de PG [@Nowacki1997], pero con trader. 

```{r}
## PG approach 
library('TRADER')

temp <- getwd()

setwd(paste0(di, 'analysis/trader_chronos/caL/'))
growthAveragingALL(caL_tree, releases = NULL, m1 = 10, m2 = 10, buffer = 10, 
                   drawing = TRUE, criteria = 0.25, criteria2 = 0.50,
                   prefix = 'CA', gfun = median, length = 5, storedev = pdf) 

setwd(paste0(di, 'analysis/trader_chronos/caH/'))
growthAveragingALL(caH_tree, releases = NULL, m1 = 10, m2 = 10, buffer = 10, 
                   drawing = TRUE, criteria = 0.25, criteria2 = 0.50,
                   prefix = 'CA', gfun = median, length = 5, storedev = pdf) 

setwd(paste0(di, 'analysis/trader_chronos/sj/'))
growthAveragingALL(sj_tree, releases = NULL, m1 = 10, m2 = 10, buffer = 10, 
                   drawing = TRUE, criteria = 0.25, criteria2 = 0.50,
                   prefix = 'SJ', gfun = median, length = 5, storedev = pdf) 
```


## Black & Abrams 2003
Análisis de boundary-line [@Black2003]
```{r}
setwd(paste0(di, 'analysis/trader_chronos/boundary_caL/'))

boundaryLineALL(caL_tree, releases = NULL, m1 = 10, m2 = 10,
                boundary = NULL, buffer = 10,
                criteria = 0.2, criteria2 = 0.5, segment = 0.5, segment2 = 0.5, 
                prefix = 'CA', drawing = TRUE, gfun = median, length = 5, 
                notop = 10, notop2 = 10, storedev = pdf) 

setwd(paste0(di, 'analysis/trader_chronos/boundary_caH/'))
boundaryLineALL(caH_tree, releases = NULL, m1 = 10, m2 = 10,
                boundary = NULL, buffer = 10,
                criteria = 0.2, criteria2 = 0.5, segment = 0.5, segment2 = 0.5, 
                prefix = 'CA', drawing = TRUE, gfun = median, length = 5, 
                notop = 10, notop2 = 10, storedev = pdf) 

setwd(paste0(di, 'analysis/trader_chronos/boundary_sj/'))
boundaryLineALL(sj_tree, releases = NULL, m1 = 10, m2 = 10,
                boundary = NULL, buffer = 10,
                criteria = 0.2, criteria2 = 0.5, segment = 0.5, segment2 = 0.5, 
                prefix = 'SJ', drawing = TRUE, gfun = median, length = 5, 
                notop = 10, notop2 = 10, storedev = pdf) 

```

## Fraver & White 2005
Análisis de absolute increase [@Fraver2005]
```{r}
setwd(paste0(di, 'analysis/trader_chronos/absoluteI_caL/'))
absoluteIncreaseALL(caL_tree, abs = NULL, m1 = 10, m2 = 10,
                buffer = 10, prefix = 'CA', drawing = TRUE, gfun = median,
                length = 5, storedev = pdf) 

setwd(paste0(di, 'analysis/trader_chronos/absoluteI_caH/'))
absoluteIncreaseALL(caH_tree, abs = NULL, m1 = 10, m2 = 10,
                buffer = 10, prefix = 'CA', drawing = TRUE, gfun = median,
                length = 5, storedev = pdf) 

setwd(paste0(di, 'analysis/trader_chronos/absoluteI_sj/'))
absoluteIncreaseALL(sj_tree, abs = NULL, m1 = 10, m2 = 10,
                buffer = 10, prefix = 'SJ', drawing = TRUE, gfun = median,
                length = 5, storedev = pdf)

setwd(temp)
```


```{r}

# SJ 
pg <- read.csv(paste0(di, 'analysis/trader_chronos/sj/', 'SJ_releases_years_total.csv'), header=TRUE) %>% 
  mutate(metodo = 'PG')
bl <- read.csv(paste0(di, 'analysis/trader_chronos/boundary_sj/', 'SJ_releases_years_total.csv'), header=TRUE) %>% 
  mutate(metodo = 'BL')
ai <- read.csv(paste0(di, 'analysis/trader_chronos/absoluteI_sj/', 'SJ_releases_years_total.csv'), header=TRUE) %>% 
  mutate(metodo = 'AI')

dist_sj <- pg %>% bind_rows(bl) %>% bind_rows(ai) %>% mutate(site = 'sj')

# caL 
pg <- read.csv(paste0(di, 'analysis/trader_chronos/caL/', 'CA_releases_years_total.csv'), header=TRUE) %>% 
  mutate(metodo = 'PG')
bl <- read.csv(paste0(di, 'analysis/trader_chronos/boundary_caL/', 'CA_releases_years_total.csv'), header=TRUE) %>% 
  mutate(metodo = 'BL')
ai <- read.csv(paste0(di, 'analysis/trader_chronos/absoluteI_caL/', 'CA_releases_years_total.csv'), header=TRUE) %>% 
  mutate(metodo = 'AI')

dist_caL <- pg %>% bind_rows(bl) %>% bind_rows(ai) %>% mutate(site = 'caL')


# caH 
pg <- read.csv(paste0(di, 'analysis/trader_chronos/caH/', 'CA_releases_years_total.csv'), header=TRUE) %>% 
  mutate(metodo = 'PG')
bl <- read.csv(paste0(di, 'analysis/trader_chronos/boundary_caH/', 'CA_releases_years_total.csv'), header=TRUE) %>% 
  mutate(metodo = 'BL')
ai <- read.csv(paste0(di, 'analysis/trader_chronos/absoluteI_caH/', 'CA_releases_years_total.csv'), header=TRUE) %>% 
  mutate(metodo = 'AI')

dist_caH <- pg %>% bind_rows(bl) %>% bind_rows(ai) %>% mutate(site = 'caH')

disturbances <- dist_caH %>% bind_rows(dist_caL, dist_sj)
```

```{r}
disturbances %>% ggplot(aes(x=AllReleasesYear, y=AllReleasesFreq)) + 
  geom_point(aes(shape=metodo, size=AllReleasesFreq)) + scale_shape(solid = FALSE) +
  facet_wrap(~site, nrow = 3) +
  theme_bw()
```

```{r}
disturbances %>% ggplot(aes(x=AllReleasesYear, y=metodo)) + 
  geom_point(aes(size=AllReleasesFreq), shape=21, colour = "black")+
  facet_wrap(~site, nrow = 3) +
  theme_bw()
```


# References 














  
  